---
title: "Meeting 13 Notes"
author: "Matthew Aiello-Lammens"
date: "5/6/2020"
output: html_document
---

# Continuing our exploration of ANOVA

## Last week - one-way anova

One response variable, one predictor variable with multiple levels
(recall, if we had one predictor variable with only two levels, could use t-test)

# Nested ANOVA

In nested ANOVA, we are accounting for potential non-independence among measurements.

## Example - Density of sea urchins influence grazing



```{r}
algae_density <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter11/Data/andrew.csv")
```

```{r}
summary(algae_density)
```

We need to convert PATCH and QUAD to factors

```{r}
algae_density$PATCH <- as.factor(algae_density$PATCH)
algae_density$QUAD <- as.factor(algae_density$QUAD)
```

```{r}
summary(algae_density)
```



First, let's do some visualizations

```{r}
library(ggplot2)

ggplot(data = algae_density) +
  geom_boxplot(aes(x = TREAT, y = ALGAE))

```

Let's incorporate the PATCHES

```{r}
ggplot(data = algae_density) +
  geom_boxplot(aes(x = TREAT, y = ALGAE, fill = PATCH))
```

Let's run an ANOVA

```{r}
algae_density_aov_nonest <- aov(data = algae_density, ALGAE ~ TREAT)
summary(algae_density_aov_nonest)
```

Let's redo the ANOVA, accounting for the nested structure

```{r}
algae_density_aov_nest <- aov(data = algae_density, ALGAE ~ TREAT + Error(PATCH))
summary(algae_density_aov_nest)
```


# Factorial (two-way) ANOVA

## Limpets, density, season -> Eggs

```{r}
limpets <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter12/Data/quinn.csv")

summary(limpets)
```

Make density into a factor

```{r}
limpets$DENSITY <- as.factor(limpets$DENSITY)

summary(limpets)
```

Some visualization

```{r}
ggplot(data = limpets) +
  geom_boxplot(aes(x = DENSITY, y = EGGS))
```

```{r}
ggplot(data = limpets) + 
  geom_boxplot(aes(x = SEASON, y = EGGS))
```

Let's look at both SEASON and DENSITY at the same time

```{r}
ggplot(data = limpets) +
  geom_boxplot(aes(x = DENSITY, y = EGGS, fill = SEASON))
```

***

This stufff added later!!!

```{r}
limpets_lm_simp <- lm(data = limpets, formula = EGGS ~ SEASON)
summary(limpets_lm_simp)

1.8398 + (-0.7360)
```








***

Let's run a model

```{r}
limpets_lm <- lm(data = limpets, formula = EGGS ~  SEASON + DENSITY)
anova(limpets_lm)
```

Look at the `lm` coefficients 

```{r}
summary(limpets_lm)
```

```{r}
library(dplyr)

limpets %>%
  group_by(SEASON) %>%
  summarise(SEASAON_Mean = mean(EGGS))

limpets %>%
  group_by(DENSITY) %>%
  summarise(DENSITY_Mean = mean(EGGS))

limpets %>%
  group_by(SEASON, DENSITY) %>%
  summarise(GROUP_Mean = mean(EGGS))

```

```{r}
library(effects)
plot(allEffects(limpets_lm))
```


```{r}
limpets_lm_int <- lm(data = limpets, EGGS ~ SEASON * DENSITY)
anova(limpets_lm_int)
summary(limpets_lm_int)
```

```{r}
plot(allEffects(limpets_lm_int))
```


## Another 2-way example

Same setup as above, but deeper into the water.

```{r}
limpets1 <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter12/Data/quinn1.csv")

summary(limpets1)
```

Convert density into a factor again

```{r}
limpets1$DENSITY <- as.factor(limpets1$DENSITY)
```


```{r}
summary(limpets1)
```

Visualization

```{r}
ggplot(data = limpets1) +
  geom_boxplot(aes(x = DENSITY, y = EGGS, fill = SEASON))
```



Run our ANOVA

```{r}
limpets1_lm <- lm(data = limpets1, EGGS ~ SEASON * DENSITY)

anova(limpets1_lm)
```

Visualize the effects with our effects plot

```{r}
plot(allEffects(limpets1_lm))
```

