---
title: "Meeting 2 - In Class Notes"
author: "Matt"
date: "2/5/2020"
output: html_document
---

# Variables and objects

Define some variables

```{r}
my_var <- 8
```

Let's make another variable

```{r}
my_var2 <- 10
```

## Make a vector

Make our first vector

```{r}
my_vect <- c(my_var, my_var2)
```

Let's make some other vectors

```{r}
v1 <- c(10, 2, 8, 7, 11, 15)

pets <- c("cat", "dog", "rabbit", "pig")
```

Make vectors of numbers based on a sequence of 
numbers

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

#### Challenge

1. Look up the help for the `seq` function, and use this function
to make a vector from 1 to 100, that steps by increments of 5

```{r}
v4 <- seq(from = 1, to = 100, by = 5)
v4

seq(1, 100, 5)
?seq

seq(from = 100, to = 1)
```


2. Come up with a way that you would use the `length.out` argument

```{r}
seq(from = 10, to = 100, length.out = 10)
```

## Exploring variable elements

Introduce the bracket notation -- `[]`

```{r}
pets
pets <- c(pets, "snake")
pets

pets[1]
pets[3]


# getting number of elements in a sequence
pets[3:4]


# getting number of elements, not in a sequenc
pets[c(1,4)]
pets
```


### Fun with vectors

```{r}
mixed_vect <- c(1, 7, "cat")
mixed_vect

as.numeric(mixed_vect[1]) + 5
```


## Working with matrices

Matrices are 2-dimensional

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
my_mat
```

##### Challenge

Break down this function, and describe each argument that is going on.


```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
matrix(data = 1:9, nrow = 3)
```

## Indexing matrices

```{r}
my_mat <- matrix(data = 1:9, nrow = 3, byrow = TRUE)
my_mat
```

```{r}
my_mat[1]
my_mat[3]
my_mat[6]

# always goes by row, column
my_mat[1,1]
my_mat[3,2]
```

```{r}
my_mat[1:2, 3]
```

```{r}
my_mat[c(1,3), 3]
```

```{r}
my_mat[c(1,3), c(1,3)]
```



```{r}
my_mat[1:2, ]
```

## Using internal functions on matrices

Make a new matrix

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
mat1
```

Apply a function to this matrix

```{r}
mean(mat1)
```

```{r}
sd(mat1)
```

calculate row and column means

```{r}
rowMeans(mat1)
```

```{r}
colMeans(mat1)
```

### `apply` function

```{r}
?apply
apply(X = mat1, MARGIN = 2, FUN = mean)
apply(X = mat1, MARGIN = 2, FUN = sd)
apply(X = mat1, MARGIN = 2, FUN = min)
apply(X = mat1, MARGIN = 2, FUN = max)

apply(X = mat1, MARGIN = 1, FUN = sd)
apply(X = mat1, MARGIN = 1, FUN = min)
mat1
```

# Data frames

*Very* special kinds of 2-d arrays

```{r}
data(iris)
head(iris)
```


```{r}
tail(iris)
```


```{r}
summary(iris)
```

### Aside: differences between characters and factors

```{r}
iris_new <- iris

iris_new$species_char <- as.character(iris_new$Species)


summary(iris_new)
```

### Introduction to navigating a data.frame

Treat a data.frame like a matrix

```{r}
iris[5, 3]
```

As a matrix, get petal length values

```{r}
iris[ , 3]
```

Just the first fifty petal length values

```{r}
iris[1:50, 3]
```

**For data.frames** I can also use the `$` operator

```{r}
iris$Petal.Length

iris$Petal.Length[5]

```

#### Challenge

What will you see when you execute the following:

```{r}
head( iris[c("Species", "Sepal.Width", "Petal.Length")] )
```


# Data Visualization

Get a new package called ggplot2

```{r, eval=FALSE}
install.packages("ggplot2")
```

```{r}
library(ggplot2)
```


### Visualizing the measurements of a single variable

```{r}
ggplot(data = iris, aes(x = Petal.Length)) +
  geom_histogram()
```

### x-y scatter plot

```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Petal.Length)) +
  geom_point()
```


```{r}
ggplot(data = iris, aes(x = Sepal.Length, y = Petal.Length,
                        colour = Species)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  xlab("Sepal Length (cm)") +
  ylab("Petal Length (cm)") +
  theme_minimal()
```



### Boxplots

```{r}
ggplot(data = iris, aes(x = Species, y = Petal.Length,
                        fill = Species)) +
  geom_boxplot()

ggplot(data = iris, aes(x = Species, y = Petal.Length,
                        fill = Species)) +
  geom_violin()

```


#### Return to histograms

```{r}
ggplot(data = iris, aes(x = Petal.Length, fill = Species)) +
  #geom_histogram(position = "dodge")
  geom_density(alpha = 0.3)
```

