---
title: "Meeting 12 - In class notes"
author: "Matthew Aiello-Lammens"
date: "4/29/2020"
output: html_document
---

# Linear Regression - Recap

* predictor variables (independent) -> continuous **numeric** value!
* response variable (dependent) -> **numeric** values!

Visualization -> x-y scatter plot

# Analysis of Variance (ANOVA)

* predictor variables (independent) -> categorical values (in R, they are called **factors**)
* response variable (dependent) -> **numeric** values

## Iris data set as example

```{r}
data(iris)
summary(iris)
```

If we compare just two of the three species, we can use a t-test

If we wanted to compare all three, one approach is to compare the three pair-wise options
* setosa - versicolor
* versicolor - virgnica
* setosa - virginica

In determining significant differences between these, we would by default apply a p < 0.05 cut off for rejecting our null hypothesis.

If I tell you, you have a 0.05 chance of flipping a coin and getting a "heads", would you have a better chance of getting at least one head if you flipped three coins?

```{r}
1 - pbinom(q = 0, size = 3, prob = 0.05)

```

How do we deal with multiple comparison problems?

We use ANOVA

```{r}
anova(lm(data = iris, Petal.Length ~ Species))
```

```{r}
TukeyHSD(aov(data =iris, Petal.Length ~ Species))
```




```{r}
library(ggplot2)

ggplot(data = NULL, aes(x = c(0, 10))) +
  stat_function(fun = df, args = list(df1 = 2, df2 = 147))
```

## Fixed effects vs random effects

* Fixed effects - a predictor variable that you have reason to suspect will have an effect; in an experiment, this is the thing you manipulate
    * Ex. does water effect plant growth? The categories of watering are fixed effect
* Random effects - a random selection of a large set of categories
    * Ex. different locations that were randomly chosen to examine percent of invasive species

For this class, we'll always work assuming fixed effects.

Mathematically - Null Hypothesis for:
* fixed = all group means are the same
* random = all group variances are the same

# One-way ANOVA 

one-way = one predictor variable

## data from Medley and Clements (Example 10A in Logan) 

What is the influence of zinc contamination on diversity of diatom species. Measured diversity shannon's index of diversity, H'.

get our data

```{r}
diatoms <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter10/Data/medley.csv")
```

look at data

```{r}
head(diatoms)
```

Visualize these data

```{r}
ggplot(data = diatoms, aes(x = ZINC, y = DIVERSITY)) +
  geom_boxplot()
```

Alternative visualization

```{r}
ggplot(data = diatoms, aes(x = ZINC, y = DIVERSITY)) +
  geom_violin()
```

Let's perform an ANOVA

Alternative ANOVA commands / functions

```{r}
table(diatoms$ZINC)
summary(aov(data = diatoms, DIVERSITY ~ ZINC))
```


Let's do this again, but use the `lm` function

```{r}
diatom_lm <- lm(data = diatoms, DIVERSITY ~ ZINC)

# Look at ANOVA results
anova(diatom_lm)
```

We can use the `summary` function on our diatom_lm object

```{r}
summary(diatom_lm)
```

We can use our model diagnostics here too

```{r}
plot(diatom_lm)
```


## Post-hoc comparisons

Tukey's Honest Significant Difference comparison test (HSD)

```{r}
TukeyHSD(aov(data = diatoms, DIVERSITY ~ ZINC))
```

```{r}
ggplot(data = NULL, aes(x = diatom_lm$residuals)) +
  geom_histogram()


ggplot(data = NULL, aes(x = diatoms$DIVERSITY)) +
  geom_histogram()

```

```{r}
ggplot(data = diatom_lm, aes(sample = DIVERSITY)) +
  geom_qq() + 
  geom_qq_line()
```




