---
title: "Introduction to Probability"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

## Section Goals

* Introduce some basic principles of probability, including conecpts of **independence** and **conditional probability**
* Learn to simulate some basic probabilistic events
* Apply R data management skills to calculate frequencies and probabilities of events using data sets


***

## Meeting preliminaries

We'll be working with the Portal Survey data this week. 
We must first load these data into our work environment.
Also, we'll be focusing only on the Rodent taxa, so let's filter the data to include only rodents

```{r}
library(tidyverse)

surveys <- read_csv("../data/portal_data_joined.csv")
surveys <-
  surveys %>%
  filter(taxa == "Rodent")
```



## Probability

> *We usually talk about probabilities in terms of events; the probability of event A occurring is written P(A). Probabilities can be between zero and one; if P(A) equals zero, then the event is impossible; if P(A) equals one, then the event is certain.* - Quinn and Keough 2002

### Frequentist statistics 

The relative frequency of an event over the long term, after many trials.

E.g., throwing a coin. What's the probability of a 'heads'? After one trial, two trials, ..., 1000 trials.

Here's an example of throwing a single coin.

```{r}
(head <- runif(1) > 0.5)
```


### Bayesian statistics

For now, all I will say is that there are other perspectives in statistics. 
A key difference between Frequentist and Bayesian statistics is that the latter allows for you to incorporate **prior knowledge**.

#### Flipping a coin - frequentist vs bayesian perspective

You find a coin on the ground, and want to know the probability that you will get a heads if you flip the coin. 
A frequentist approach would flip the coin many times and assume the probability of getting a heads is approximately the **frequency** of heads.
A Bayesian approach would assume that the probability is 0.5 based on a naive guess, then collect data (i.e., flip the coin) and update the *prior* probability using the newly acquired data.

***


## Properties of probability

* a **random process** gives rise to an **outcome**
    * a coin-flip is a random process
    * capturing a *Peromyscus* in the Portal traps
* 0 $\le$ P(*some event*) $\le$ 1
* $\hat{p}_n$ is the proportion of outcomes out of the total number of **trials** run
    * If we flip a coin 100 times, and 52 times we got heads, $\hat{p}_{100} = \frac{52}{100}$
    * Out of `r nrow(surveys)` total rodent captures, `r sum(surveys$genus == "Peromyscus")` were *Peromyscus*, thus the probability of capturing a *Peromyscus* is `r round(sum(surveys$genus == "Peromyscus") / nrow(surveys), 2)`

### Law of Large Numbers

> *The tendency of $\hat{p}_n$ to stabilize around $p$ [as $n$ increases] is described by the **Law of Large Numbers**.* - Vu and Harrington 2020, p. 92

#### Example of law of large numbers

Let's start by calculating the frequency of each of the genera captures in the Portal traps. 
These frequency values can be interpreted as probability values of catching a particular genus in a trap.

```{r}
genus_frequencies <-
  surveys %>%
  group_by(genus) %>%
  count()

genus_frequencies$freq <- round(genus_frequencies$n / sum(genus_frequencies$n), 4)

knitr::kable(genus_frequencies)
```

The probabilities above are based on **full information**. 
What happens if we only has partial information?

First, how can we get a subset of the data?

```{r}
# Pick the number of rows to select
rows_to_keep <- sample(1:nrow(surveys), size = 100, replace = FALSE)

surveys_subset <- surveys[rows_to_keep, ]
```

Next let's calculate the frequencies / probabilities with the subset of data.

```{r}
gen_freq_sub <-
  surveys_subset %>%
  group_by(genus) %>%
  count()

gen_freq_sub$freq <- round(gen_freq_sub$n / sum(gen_freq_sub$n), 4)

gen_freq_sub

```

Next, let's play with the numbers of samples, and see what happens to the frequencies compared to the *real* frequencies.

```{r}
# Pick the number of rows to select
rows_to_keep <- sample(1:nrow(surveys), size = 1000, replace = FALSE)

surveys_subset <- surveys[rows_to_keep, ]

gen_freq_sub <-
  surveys_subset %>%
  group_by(genus) %>%
  count()

gen_freq_sub$freq <- round(gen_freq_sub$n / sum(gen_freq_sub$n), 4)

gen_freq_sub

```


#### Aside: sampling with vs. without replacement

Above, we set `replace = FALSE` in the `sample` function.
This means that any row that is selected, can only be selected **once**. 

On the other hand, if we set `replace = TRUE`, then any row can be selected more than once *by random*. 

```{r}
# Pick the number of rows to select - WITH REPLACE = TRUE!
rows_to_keep <- sample(1:nrow(surveys), size = 1000, replace = TRUE)

surveys_subset <- surveys[rows_to_keep, ]

gen_freq_sub <-
  surveys_subset %>%
  group_by(genus) %>%
  count()

gen_freq_sub$freq <- round(gen_freq_sub$n / sum(gen_freq_sub$n), 4)

gen_freq_sub

```

With a sample size of 1000, these values should be pretty similar.
But there are reasons why you might choose `replace = TRUE`. 
For example, if you are projecting probabilities of capturing genera *in the future*, then treating past observations as a way to build generic distributions (using `replace = TRUE`) may be more appropriate.

### Specific examples of calculating probabilities

1. Probability of capturing a *Neotoma*

P(*Neotoma*) = 0.0366 = 3.6%

2. Probability of capturing a *Neotoma* OR a *Dipodomys*

These are **mutually exclusive** events. You *cannot* be both a *Neotoma* and a *Dipodomys*.

P(*Neotoma* OR *Dipodomys*) = 0.0366 + 0.4721 = 0.5087

3. Probability of capturing something **other than** *Onychomys*

This is called the **complement** of *Onychomys*

P(not *Onychomys*) = 1 - 0.0954 = 0.9046

***

#### Challenge - more with probability

1. What is the probability of capturing a **female** rodent?

2. What is the probability of capturing a female *Dipodomys*?

P(female AND *Dipodomys*)

3. What is the probability of capturing a female rodent OR a *Dipodomys*?

*Note* - these are non-independent events. You need to use the **General Addition Rule**.

$$
P(A \text{ or } B) = P(A) + P(B) - P(A \text{ and } B)
$$


### Independence

> *Two processes are **independent** if knowing the outcome of one provides* no useful information *about the outcome of the other.* - Vu and Harrington 2020, p. 101

***

#### Challenge 

* Is flipping two coins different from flipping one, then the other? 
* You flip 99 heads. What is the probability you will flip a head on the 100th toss?
* You catch a *Neotoma* in a trap in one of the Portal plots. What is the probability that you will catch a *Neotoma* in a trap in one of the other Portal plots?


***

### Multiplication Rule for Independent Processes

From *Vu and Harrington 2020, p. 103*

> If A and B represent events from two different and independent processes, then the probability that both A and B occur is given by:

$$
P(A \text{ and } B) = P(A) \times P(B)
$$

> Similarly, if there are $k$ events $A_1$, ..., $A_k$ from $k$ independent processes, then the probability they all occur is:

$$
P(A_1)P(A_2)...P(A_k)
$$

***

#### Challenge - probability of capturing 100 *Dipodomys* in a row vs capturing 100 *Chaetodipus* in a row.

First recall how we calculated the frequencies / probabilities of capturing each of the different genera of rodents. Then use the multiplication rule to calculate the two probabilities above.

```{r}
genus_frequencies <-
  surveys %>%
  group_by(genus) %>%
  count()

genus_frequencies$freq <- round(genus_frequencies$n / sum(genus_frequencies$n), 4)

knitr::kable(genus_frequencies)
```

***

### Conditional Probability

#### Conditional Probability: Intuition

Consider height in the US population.

What is the probability that a randomly selected individual in the population is taller than 6 feet, 4 inches?

* Suppose you learn that the individual is a professional basketball player. Does this change the probability that the individual is taller than 6 feet, 4 inches?

### Conditional Probability: Concept

The **conditional probability** of an event $A$, given a second event $B$, is the probability of $A$ happening, knowing that the event $B$ has happened.

* This conditional probability is denoted $P(A|B)$

What is the probability of trapping a *Dipodomys merriami*?

Let $A$ = capture a *Dipodomys merriami*

```{r}
species_frequencies <-
  surveys %>%
  group_by(genus, species) %>%
  count()

species_frequencies <-
  species_frequencies %>%
  mutate(freq = round(n / sum(species_frequencies$n), 4))

```

What is the probability of trapping a *Dipodomys merriami*, conditional (or *given*) that you captures a *Dipodomys*?

Again, let $A$ = capture a *Dipodomys merriami*, and now $B$ = capture a *Dipodomys*.

```{r}
dipo_frequencies <-
  surveys %>%
  filter(genus == "Dipodomys") %>%
  group_by(species) %>%
  count()

dipo_frequencies <-
  dipo_frequencies %>%
  mutate(freq = round(n / sum(dipo_frequencies$n), 4))

dipo_frequencies
```


#### Conditional Probability: Formal Definition

As long as $P(B) > 0$,
\[P(A|B) = \dfrac{P(A \cap B)}{P(B)}. \]

From the definition,
\begin{align*}
P(A|B) =& \dfrac{P(\text{capture a Dipodomys merriami AND capture a Dipodomys})}{P(\text{capture a Dipodomys})} \\
=& \dfrac{0.3094}{0.4721} = 0.6554
\end{align*}

### Independence, Again...

A consequence of the definition of conditional probability:

* If $P(A|B) = P(A)$, then $A$ and $B$ are independent; knowing $B$ offers no information about whether $A$ occurred.


### General Multiplication Rule

If $A$ and $B$ represent two outcomes or events, then
\[P(A \cap B) = P(A|B)P(B).\]

This follows from rearranging the definition of conditional probability:
\[P(A|B) = \frac{P(A \cap B)}{P(B)} \rightarrow P(A|B)P(B) = P(A \cap B)\]

Unlike the previously mentioned multiplication rule, this is valid for events that might not be independent.

### Bayes' Theorem, aka Bayes' Rule

Bayes' Thorem (simplest form):

$$ P(A|B) = \frac{P(B|A)P(A)}{P(B)}$$


Follows directly from the definition of conditional probability by noting that $P(A) P(B|A)$ equals $P(A \text{ and } B)$:

\[P(A|B) =  \frac{P(A \cap B)}{P(B)} = \frac{P(B|A)P(A)}{P(B)}\]

Foundation of Bayesian statistics:

\[P(\text{assumptions}|\text{data}) =  \frac{P(\text{data}|\text{assumptions})P(\text{assumptions})}{P(\text{data})}\]



