---
title: "Meeting 7 - Introduction to Biostats and R Programming"
author: "Matthew E. Aiello-Lammens"
date: '2019-03-13'
output:
  html_document:
    code_folding: hide
    toc: yes
  pdf_document:
    toc: yes
---

# Class goals

* Learn how to develop a research question - **DONE**
* Become proficient using the R statistical programming language
* Learn how to explore data collected during a research project
* Learn and master the standard statistical analyses used in scientific studies
* Understand the connections between statistical analyses and study design
* Become familiar with resources to explore advanced analysis methods

# Class format

* Some lecture - mostly about concepts
* Lots of **live coding** examples - you should follow along, but don't have to
* Challenges - some will be short, some long. Some you work on alone, some with a partner or in groups

# Hypothesis testing and the scientific method

### What is a theory?

A proposed explanation for some event or process that is based on the available evidence related to that event.
We also call this a **model**. *Models don't necessarily mean **math**.*

### What is a hypothesis?

A prediction based on our model (or theory). 
For a hypothesis to be considered **scientific** it must be **falsifable**.

## Scientific method

Traditional view:

1. Develop a hypothesis 
2. Make a prediction based on hypothesis 
3. Do an experiment to test the prediction
4. Analyze the data from our experiment
5. Based on results, make an inference about our hypothesis - do the data support or reject our prediction?
6. Repeat as necessary to acquire knowledge

Checkout [Understanding Science](http://undsci.berkeley.edu/article/0_0_0/howscienceworks_02)

### Analysis is iterative

The process of analyzing data is often iterative. 
That means, we may start looking at our data, applying analysis techniques that might *not* be the techniques we use in our final report (i.e., our reports and manuscripts). 
Don't be afraid to start exploring your data, and applying analyses, before you think you're ready.

# Introduction to R and RStudio

## Why do I have to learn to program in R?

In short, it's "industry standard". 
But the longer answer has to do with replicability of data analyses.
Documenting your analysis in R, using R scripts or Rmd files allows you to re-run your analysis whenever the need arises, and to share your analysis workflow with others.

## Getting started

### Difference between R and RStudio

In this class, we will be working primarily in RStudio.
So what is the difference between R and RStudio?
R is both a programming language (specifically a statsitical analysis programming language), while RStudio is an **Integrated Development Environment**, or **IDE** for short. 
RStudio offers a number of features, mostly related to visual presentation of information, that make writing and working with R code easier.

### Overall layout

* Four panels, each has valuable information

### File management

* Setup a `ENS-623-Research-Stats` directory
* Make a `data` directory
* Make a `scripts` directory
* In RStudio, set working directory to your new `ENS-623-Research-Stats` directory

#### Setting your working directory

Point-and-click method - Use 'Session' > 'Set Working Directory' > 'Choose Directory'.

Using the R Console:

```{r, eval=FALSE}
setwd("/Users/maiellolammens/Dropbox/Pace/Teaching/ENS-623-Research-Stats/ENS-623-Research-Stats-SP18/")
```


### Getting help

* **Help** panel (lower right corner)
* `help.search`

```{r}
help.search("bar plot")
```

#### Challenge

Use the `help.search` function to search for something in statistics that you think should be in R? Did you find anything?

* I know my function - just give me the details - `?barplot`


## R as calculator

We can use R just like any other calculator.

```{r}
3 + 5
```

There's internal control for order of operations (Please Excuse My Dear Aunt Sally)

```{r}
(3 * 5) + 7
```

```{r}
3 * 5 + 7
```

#### Challenge

Write an example where adding parentheses matters.

### Internal functions

There are a ton of internal functions, and a lot of add-ons.

```{r}
sqrt(4)

abs(-5)

sqrt(-5)
```

## R script file

Use a script file for your work. It's easier to go back to and easy to document.

**Important:** within an R file, you can use the # sign to add comments. Anything written after the # is *not* interpreted when you run the code.

#### Challenge

Create a new R script file in your `scripts` directory.

### Basic file managment in R

```{r}
# What working directory am I in?
getwd()

# Move to a different director?
setwd(".")
```

#### Things to cover

* Navigating the file path
* Tab completion of file paths
* Tab completion of R commands

#### Challenge

* Try to auto-complete `fil`, what do you find?
* Use the brief help menu that comes up to find a function that starts with `file`, and describe what you think it does.

## Rmd file

Use this to integrate text and R code into the same document.
**I will expect most of your homework assignments as an Rmd file.**

*Practice with Rmd file*

## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

### Functions that are useful for understanding the different types of data structures

```
str()
class()
```

## Practice with variables

Define a variable

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Work with vars

```{r}
my_var + my_var2
```

Make a new variable

```{r}
my_var_tot <- my_var + my_var2
```

#### Challenge

Change the value of my_var2

```{r}
my_var2 <- 3
```

What is the value of `my_var_tot` now?

### Make a vector

Combining values into a vector

```{r}
# Vector of variables
my_vect <- c(my_var, my_var2)

# Numeric vector
v1 <- c(10, 2, 8, 7, 11, 15)

# Char vector
pets <- c("cat", "dog", "rabbit", "pig")
```

Making a vector of numbers in sequence

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

#### Challenge

1. Look up the help for the `seq` function, and use this to make a vector from 1 to 100, by steps of 5.
2. Come up with a way that you would use the `length.out` argument.

## Exploring variable elements

You can get specific elements from vectors and other data structures

* Introduction to the square brackets `[]`

```{r}
pets <- c("cat", "dog", "rabbit", "pig", "snake")
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```


## Working with matrices

**Review** - Why might we want 2D data?

Let's make a matrix

#### Challenge

With the people next to you, break down this function, and describe each argument.
What is the final product?

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

What does it mean to fill `byrow`?

```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

Versus

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```

#### Challenge

What is the default value for `byrow`?

### Indexing matrices

Indexing happens by **row**, **column** notation.

```{r}
my_mat <- matrix(data = 1:50, nrow = 10, byrow = TRUE)

my_mat[1,1]
my_mat[1,2]
my_mat[2,1]

my_mat[1:4, 1:3]

my_mat[c(1,3,5), ]
my_mat[ ,c(1,3,5)]
```

## Combinining internal functions with matrices

Make a "random" matrix (that isn't random in this case because of the `set.seed` function)

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

Calculate the mean of all of the data

```{r}
mean(mat1)
```

Calculate the standard deviation of all of the data

```{r}
sd(mat1)
```

Calculate row means and column means

```{r}
rowMeans(mat1)
colMeans(mat1)
```

Introduce the `apply` function

```{r}
apply(mat1, MARGIN = 1, mean)
```

## Data frames

### Iris dataset

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data(iris)
```

Let's have a look at this dataset

```{r}
head(iris)
tail(iris)
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

#### `summary` function

Let's begin by using the `summary` function to examine this dataset.
`summary` returns many of the standard statistics.
When doing data exploration, a few things you want to look at are:

* How to the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(iris)
```

#### Aside: Differences between characters and factors

Factors 'appear' to be similar to characters, but are in fact coded numerically in R. 
Think of **factors** like **categories**. 
Here's a quick example that demonstrates the difference in these two variable types that shows up when using `summary`.

```{r}
## Make a new iris dataset
iris_new <- iris

## Create a new column that treats species as a character, rather than a factor
iris_new$species_char <- as.character(iris_new$Species)

## Run summary command
summary(iris_new)
```


#### Aside: A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
I recommend checking out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use the *row* and *column* notation.

Here is the 5th row, 3rd column (Petal.Length).
**Note: We are using square brackets to index the `data.frame` and we *always* use row, column notation.**

```{r}
iris[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `Petal.Length` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
iris[ ,3]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
iris$Petal.Length
```

* Looking at specific column entry

This is another way to look at the 5th entry in the `Petal.Length` column.

```{r}
iris$Petal.Length[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**Note: here we leave the column part blank, but still add the comma.**

```{r}
iris[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 1st through 3rd columns.
**Note: we use the `:` operator to look at a range of value.**

```{r}
iris[5:10, 1:3]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(iris[2:3])
```

***

#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

## Reading in your own data

One of the most basic things you will need to do in R is read in your own data set. 
You can read in Excel files, simple text files, and even files from Google Sheets.
But the easiest type of file to read in is a comma seperated values (CSV) file. 
You can save an Excel workbook (or Numbers workbook or Google Sheet) as a CSV file by using the "Save as ..." menu item. 

For today, we're going to work with a data set related to gastropod fecundity.
We'll link to the file directly using a Dropbox link, but it's possible to download the file here - [gastropod data](https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter6/Data/ward.csv) **The download will begin when you press the hyperlink! The file is called `ward.csv`.**

### Differences in fecundity of intertidal gastropods in two different intertidal zones (Example 6A in Logan, Box 3.1 in Q&K)

Ward and Quinn (1988) investigated the differences in fecundity of *Lepsiella vinosa* in two different intertidal zones (mussel zone and littorinid zone). 

Get the data and have a quick look

```{r}
gastro <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter6/Data/ward.csv")
summary(gastro)
```




