---
title: "Meeting 2 - R Programming and Data Visualizatoin"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

# Today's Goals

* Finish introduction to R materials
* Introduce some basic programming techniques used in reproducible analysis
* Become familiar with data visualization using the `ggplot2` package

***

Before we do anything, let's make sure we are all working with a blank slate. Re-start RStudio and clear your environment if there's anything in there.

## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

### Functions that are useful for understanding the different types of data structures

```
str()
class()
```

## Practice with variables

### Variables (aka Objects)

Define a variable

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Work with vars

```{r}
my_var + my_var2
```

Make a new variable

```{r}
my_var_tot <- my_var + my_var2
```

***

#### Challenge

Change the value of my_var2

```{r}
my_var2 <- 3
```

What is the value of `my_var_tot` now?

***


### Vectors

Define a variable

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Combining values into a vector

```{r}
# Vector of variables
my_vect <- c(my_var, my_var2)
```


Here are some other examples of vectors

```{r}
# Numeric vector
v1 <- c(10, 2, 8, 7, 11, 15)

# Char vector
pets <- c("cat", "dog", "rabbit", "pig")
```

Making a vector of numbers in sequence

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

***

#### Challenge

1. Look up the help for the `seq` function, and use this to make a vector from 1 to 100, by steps of 5.
2. Come up with a way that you would use the `length.out` argument in the `seq` function.

***

### Exploring variable elements

You can get specific elements from vectors and other data structures

#### Introduction to the square brackets `[]`

Using the square bracket notation is sometimes referred to as **indexing**, or **index notation**.

```{r}
pets <- c("cat", "dog", "rabbit", "pig", "snake")
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```


### Working with matrices

**Review** - Why might we want 2D data?

Let's make a matrix

***

#### Challenge

With a partner, break down this function, and describe each argument.
What is the final product?

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

***


What does it mean to fill `byrow`?

```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

Versus

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```

***

#### Challenge

What is the default value for `byrow`?

***

### Indexing matrices

Indexing happens by **row**, **column** notation.

```{r}
my_mat <- matrix(data = 1:50, nrow = 10, byrow = TRUE)

my_mat[1,1]
my_mat[1,2]
my_mat[2,1]

my_mat[1:4, 1:3]

my_mat[c(1,3,5), ]
my_mat[ ,c(1,3,5)]
```

## Using internal functions on matrices

Make a "random" matrix (that isn't random in this case because of the `set.seed` function)

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

Calculate the mean of all of the data

```{r}
mean(mat1)
```

Calculate the standard deviation of all of the data

```{r}
sd(mat1)
```

Calculate row means and column means

```{r}
rowMeans(mat1)
colMeans(mat1)
```

Introduce the `apply` function

```{r}
apply(mat1, MARGIN = 1, mean)
```

## Data frames

### Iris dataset

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data(iris)
```

Let's have a look at this dataset

```{r}
head(iris)
tail(iris)
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

#### `summary` function

Let's begin by using the `summary` function to examine this dataset.
`summary` returns many of the standard statistics.
When doing data exploration, a few things you want to look at are:

* How do the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(iris)
```

#### Aside: Differences between characters and factors

Factors 'appear' to be similar to characters, but are in fact coded numerically in R. 
Think of **factors** like **categories**. 
Here's a quick example that demonstrates the difference in these two variable types that shows up when using `summary`.

```{r}
## Make a new iris dataset
iris_new <- iris

## Create a new column that treats species as a character, rather than a factor
iris_new$species_char <- as.character(iris_new$Species)

## Run summary command
summary(iris_new)
```


#### Aside: A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
I recommend checking out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use the *row* and *column* notation.

Here is the 5th row, 3rd column (Petal.Length).
**Note: We are using square brackets to index the `data.frame` and we *always* use row, column notation.**

```{r}
iris[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `Petal.Length` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
iris[ ,3]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
iris$Petal.Length
```

* Looking at specific column entry

This is another way to look at the 5th entry in the `Petal.Length` column.

```{r}
iris$Petal.Length[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**Note: here we leave the column part blank, but still add the comma.**

```{r}
iris[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 1st through 3rd columns.
**Note: we use the `:` operator to look at a range of value.**

```{r}
iris[5:10, 1:3]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(iris[2:3])
```

***

#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

# Data Visualization and Exploration

The first step to any data analysis is to interrogate your data by calculating some standard statistics and by visualizing it in various ways. 
We just looked at a few ways to summarize a data set, and in future classes we will look more closely at standard statistics.
Today, we'll focus on data visualization.

For the next section, we're going to continue to work with the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).

***

#### Challenge

What commands can I use if I want the first 6 elements of both the sepal width and the species columns?

***

#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

Before continuing with visualization, let's add a habitat type variable to the iris data set. We'll use this later. 
*Caveat - I made up these habitat type preferences.*

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )

iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )

head( iris_full )
tail( iris_full )
```


## Visualization using the `ggplot2` package

I'm going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved than base R (i.e., the `graphics` package).

### Aside: Installing and loading packages

First, we need to install the `ggplot2` package:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```

**NOTE: You will only have to intall a new package once, but you will need to call the `library` function at the beginning of every new R session.**

### Visualizing the measurements of a single variable

Perhaps the most common way to look at data for a single variable is a histogram.
This essentially is a bar plot, where each bar represents the number of times a value falls within a particular *bin*.

Example: let's look at the distribution of `Petal.Length` values

```{r}
ggplot(data = iris_full, aes(x = Petal.Length)) +
  geom_histogram() 
```

Let's break down this call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

***

#### Challenge - use `?geom_histogram` to determine how to change the number of bins used

```{r, echo=FALSE, eval=FALSE}
ggplot(data = iris_full, aes(x = Petal.Length)) +
  geom_histogram(bins = 100) 
```

***

### Visualizing relationships between two variables

Now, let's make an x-y scatter plot. 
Specifically, we would like to plot Sepal.Length versus Petal.Length. 

```{r plot1-ggplot}
ggplot(data = iris_full, aes( x = Sepal.Length, y = Petal.Length )) + 
  geom_point(  )
```


**THAT SEEMS SO COMPLICATED!**

It's true. The syntax for ggplot can seem pretty complicated. 
But the power of ggplot lies in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. 
For example, let's say I want to create the plot we just made, but have each species represented by a different color.

```{r plot2}
ggplot(data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) ) + 
  geom_point( )
```

Let's add more information - how about habitat type as well.

```{r}
ggplot(data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species, shape = Habitat)) + 
  geom_point(size = 2.5 ) +
  theme_bw()
```

How about if we wanted to add trend lines.

```{r}
ggplot(data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species)) + 
  geom_point(size = 2.5 ) +
  geom_smooth(method = "lm") +
  theme_bw()
```

**facets** - a way to separate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`.

```{r plot2-facets}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```


*** 

#### Challenge

1. `ggplot2` has many geometries, allowing us to make lot's of different 
types of plots. Let's make two new plots - one **boxplot** of `Petal.Length`, 
with one boxplot for each species. Use `geom_boxplot` for this.

```{r}
ggplot(data = iris_full, aes(x = Species, y = Petal.Length, fill = Species)) +
  geom_boxplot() +
  theme_bw()  
```

2. Make a histogram of Petal.width. 

```{r}
ggplot(data = iris_full, aes(x = Petal.Length)) +
  geom_histogram() +
  theme_bw()  
```


Use facets to separate the three species.

```{r}
ggplot(data = iris_full, aes(x = Petal.Length, fill = Species)) +
  geom_histogram() +
  facet_grid( Species ~ . ) +
  theme_bw()  
```

3. Second, let's make density plots of Petal.Width. Use `geom_density` and
`colour` to make different colored density lines for iris in each 
habitat type. 
**Note: the area under the curve is equal to 1.**

```{r}
ggplot(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  geom_density() +
  theme_bw()
```

4. Use histogram to plot density instead of counts.

```{r}
ggplot(data = iris_full, aes(x = Petal.Length)) +
  geom_histogram(aes(y = ..density.., fill = Species)) +
  facet_grid( Species ~ . ) +
  geom_density(aes(colour = Species)) +
  theme_bw()  
```


***

### More `ggplot2` resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.
