---
title: "Meeting 4 - Working with Data in R"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

> *Note - much of the material in this session is based on or remixed from materials in the [Data Carpentry - Data Analysis and Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/index.html) lessons, maintained by François Michonneau & Auriel Fournier. Copyright (c) Data Carpentry.

# Today's Goals

* Be able to load a data set into the R environment from a .csv file
* Practice with basic data summary and `data.frame` indexing functions
* Practice working with **factors** and **dates**
* Become familiar with the `tidyverse` "curated" list of packages
* Practice subsetting a data set using functions `select` and `filter`
* Practice calculating summaries of a data set usinig functions `group_by` and `summarize`

***

# Portal Project Data Set

We will be working with a data set from the [Portal Project](https://portal.weecology.org/), a long-term ecological research project near Portal, Arizona (southeast AZ) that was started in 1977. 
Data collected as part of this project includes observations of plants, rodents, and ants in a set of experimentally manipulated study plots. 

### Downloading the data

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from [figshare](https://figshare.com/), a website for storing and archiving research products.
Here's a direct link to the figshare page for these data - [https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10](https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10) - but we won't be downloading these data manually.

Inside the `download.file` command, the first entry is a character string with the
source URL ("https://ndownloader.figshare.com/files/2292169"). 
This source URL downloads a CSV file from figshare. The text after the comma
("data/portal_data_joined.csv") is the destination of the file on your local
machine. You'll need to have a folder on your machine called "data" where 
you'll download the file to. 
You should have created this folder in the first week of class.
So this command downloads a file from figshare, names 
it "portal_data_joined.csv" and adds it to a preexisting folder named "data".

```{r, eval=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

```{r, echo=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "../data/portal_data_joined.csv")
```


### Reading the data into R

The file has now been downloaded to the destination you specified, but R has not 
yet loaded the data from the file into memory. 
To do this, we are going to use the `read_csv()` function from the **`tidyverse`** package.
Note that there other ways to read data in, such as the base R `read.csv()`, but we'll use the upgraded `read_csv()` function here.

First, we need to install the **`tidyverse`** package. 
Type `install.packages("tidyverse")` straight into the console. 
It's better to write this in the console than in our script for any package, as there's no 
need to re-install packages every time we run the script. 
Then, to load the package type:

```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

Now we can use the functions from the **`tidyverse`** package. 
Let's use `read_csv()` to read the data into a data frame:

```{r, eval=FALSE,  purl=FALSE}
surveys <- read_csv("data/portal_data_joined.csv")
```

```{r, eval=TRUE, echo=FALSE,  purl=FALSE}
surveys <- read_csv("../data/portal_data_joined.csv")
```


You will see the message `Parsed with column specification`, followed by each
column name and its data type.
When you execute `read_csv` on a data file, it looks through the first 1000 rows 
of each column and guesses its data type. For example, in this data set,
`read_csv()` reads `weight` as `col_double` (a numeric data type), and `species` 
as `col_character`. You have the option to specify the data type for a column
manually by using the `col_types` argument in `read_csv`.

### Examining the `surveys` data frame

Let's use some of the functions we've learned in the past, and a few new ones, to look at the `surveys` data frame.

First, let's open the data frame in a viewer.

```{r}
View(surveys)
```

Next, let's use the `summary` function.

```{r}
summary(surveys)
```

How about a few other functions?

***

#### Challenge - looking a data frames

Try using each of the functions below and identifying what they do.

```{r, eval = FALSE}
dim(surveys)
nrow(surveys)
ncol(surveys)
names(surveys)
rownames(surveys)
str(surveys)

```

***

