---
title: "Meeting 4 - Research Questions and Introduction to Probability"
author: "Matthew E. Aiello-Lammens"
date: '2020-02-19'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Review the process of developing research questions
* Introduce some basic principles of probability
* Learn to simulate some basic probabilistic events
* Apply R programming skills to simulate the process of flipping a coin until you get 100 heads


***

# Asking research questions

Below are some quotes and thoughts from the Karban et al. reading I thought important to touch on in discussion.

> ...at this statge in most students' careers, many topics sound equally interesting, so this forced focus is difficult or even painful.

How many of you feel this way?

> For most jobs ... search committees will want to see strong records of research and publication even if you won't be expected to conduct research or publish a lot on the job.

What research gives us (from Box 1) - 

* Multiple perspectives
* Reasoning and critical thinking
* Writing and communication

***

#### Challenge 

Write down some short-, mid-, and long-term goals related to environmental science.

***

> So you're looking for questions that are specific yet general and novel yet relevant to your goals. You could fret about this for years.

Personal note - I did not defend my dissertation *proposal* until the end of my third year as a PhD student. I fretted for a few years. 

However, for your MS in Environmental Science, novelty is not necessarily as important as it is for a PhD in Ecology (at least in our program).

> ... it is not realistic to expect yourself to sit at your desk and conjure up the perfect study ...

Get out and do things. Experiment. Mess around with data. Write grant applications before you're ready. It will all stimulate thinking and creativity.

> Fieldwork is hard business, and many of the factors associated with failure or success are beyond your control.

Some field projects fail. Some don't yield interesting results. At the MS level, we're more interested in refining your process. **Focus on the process.**

***

#### Challenge

Considering formulating a research question, are you more drawn to asking some specific question first or working in a particular system?

***

> Don't assume questions have been answered just because they seem obvious.

> Be willing to hang with the dumb ideas that you will inevitably come up with, because the really great ideas stand on the shoulders of the dumb ones.


***

# Probability

> *We usually talk about probabilities in terms of events; the probability of event A occurring is written P(A). Probabilities can be between zero and one; if P(A) equals zero, then the event is impossible; if P(A) equals one, then the event is certain.* - Quinn and Keough 2002

## Frequentist statistics 

The relative frequency of an event over the long term, after many trials.

E.g., throwing a coin. What's the probability of a 'heads'? After one trial, two trials, ..., 1000 trials.

```{r}
(head <- runif(1) > 0.5)
```


## Bayesian statistics

For now, all I will say is that there are other perspectives in statistics. 
A key difference between Frequentist and Bayesian statistics is that the latter allows for you to incorporate **prior knowledge**.

### Flipping a coin - frequentist vs bayesian perspective

You find a coin on the ground, and want to know the probability that you will get a heads if you flip the coin. 
A frequentist approach would flip the coin many times and assume the probability of getting a heads is approximately the frequency of heads.
A Bayesian approach would assume that the probability is 0.5 based on past experience, then collect data (i.e., flip the coin) and update the *prior* probability using the newly acquired data.

***


# Properties of probability

* a **random process** gives rise to an **outcome**
    * a coin-flip is a random process
* 0 $\le$ P(*some event*) $\le$ 1
* $\hat{p}_n$ is the proportion of outcomes out of the total number of **trials** run
    * If we flip a coin 100 times, and 52 times we got heads, $\hat{p}_{100} = \frac{52}{100}$

## Law of Large Numbers

> *As more observations are collected, the proportion of $\hat{p}_n$ of occurrences with a particular outcome converges to the probability of $p$ of that outcome.* - Diez et al. 2016, p. 77

## Example of law of large numbers

Let's simulate a coin flip. To do this, we'll draw a single uniform random number between 0 and 1.

```{r}
(flip <- runif(1))
```

Let's define all values between 0 and 0.5 as being a "heads".
In order to *define* the flip as heads or tails, we'll need to use an `if` statement.

```{r}
if(flip < 0.5){
  coin_flip <- "head"
} else {
  coin_flip <- "tail"
}
```

Let's simulate the coin flip 100 times, and count the number of heads we get.

```{r}
# start a vector to store all of the flips
all_flips <- c()

for(x in 1:100){
  # "flip" the coin
  flip <- runif(1)
  
  # Identify the flip as heads or tails
  if(flip < 0.5){
    coin_flip <- "head"
  } else {
    coin_flip <- "tail"
  }
  
  # add this flip to the vector of all_flips
  all_flips <- c(all_flips, coin_flip)
  
}


```


Now, let's count the number of heads.

```{r}
all_flips == "head"
```

```{r}
sum(all_flips == "head")
```

Calculate the probability of a heads

```{r}
sum(all_flips == "head") / length(all_flips)
```

How close to 0.5 is this?

***

#### Challenge

Repeat the steps above until you get to a number of flips to make that consistently results in values *very* close to 0.5.

***

# R programming recap

## Challenge: Number of coin flips it takes to get to 100 heads

Imagine you are flipping a fair coin (i.e., there is a 0.5 chance of getting a heads and a 0.5 chance of getting a tails).
You want to continue flipping the coin *until* you get 100 heads.
**Write a script that simulates coin flips, and counts the number of times you need to flip the coin until you get 100 heads.**
Note that since the coin flip is a random process, it will not necessarily take you the same number of flips if you re-run the script.

### Pseudo code

As we started last meeting, you might want to begin by writing some pseudo code

1. Begin by writing down the main things you need to do

```
- Flip a coin
- Determine if the coin is "heads" or "tails"
- Repeat the flip until you get 100 heads
- Tally the number of heads and the number of total flips
```

2. Start putting these things in order

3. Add in key components of the flow-control


*** 

# Work on problem set 3

If there is extra time in class this week, we will begin working on problem set 3 as a group.