---
title: "Meeting 5 - Probability"
author: "Matthew E. Aiello-Lammens"
date: '2020-02-26'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Review the statistics mean and variance / standard deviation
* Introduce basic principles of probability
* Introduce procedures to code the simulation of a random event - i.e., getting a particular bag of M&Ms

***



# Probability through M&Ms

Open your bag(s) and count how many M&Ms you got. Also count how many of each color.

Make a new variable called `tot_mm` and assign it the number of MMs in your bag.

```{r}
tot_mm <- 60
```

## Average number of MMs in a bag

Calculate the mean number of MMs in a bag.

```{r}
tot_mm_pop_sample <- c(60, 58, 57, 63, 61, 60, 59)
```

Mean:

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$

Using the `sum` function.

```{r}
(mean_mm_1 <- sum(tot_mm_pop_sample) / length(tot_mm_pop_sample))
```

Using the `mean` function.

```{r}
(mean_mm_2 <- mean(tot_mm_pop_sample))
```

## Quantify the variation in the number of MMs in a bag

* `var` function
* `sd` function

**What do we mean by variation?**

In general, we mean something along the lines of 'the amount of variation around some mean value'.

**How might we quantify this?**

* Add up the distances from the mean. 

First calculate distances from the mean for each bag.

```{r}
(dist_from_mean <- tot_mm_pop_sample - mean_mm_1)
```

Next calculate the total distances. We could also call this the **sums of distances**

```{r}
(sum_dist_from_mean <- sum(dist_from_mean))
```

What information does this give us, and why might it not be useful?

* Add up the *absolute* distances from the mean. 

First calculate the *absolute* distances from the mean.

```{r}
(absdist_from_mean <- abs(tot_mm_pop_sample - mean_mm_1))
```

Next add up the total *absolute* distances from the mean.

```{r}
(sum_absdist_from_mean <- sum(absdist_from_mean))
```

While the absolute distance from the mean does give us a reasonable measure of variability, because of specific mathematical properties, it's more convenient to work with *squared* distances from the mean, leading to the measures of **variance** and **standard deviation** we are more familiar with.

* Add up the *squared* distances from the mean.

***

#### Challenge

Calculate *squared* distances from the mean, and sum them to determine the total squared deviation.

***

**Variance** is the average squared distance from the mean -- need to divide by $n$ (actually, we divide by $n-1$, but we're not going to go into why just yet.)

### Sample Variance

$$
var = \frac{\sum_{i=1}^n{(\bar{x} - x_i)^2}}{n-1}
$$


**Standard deviation** can be thought of as the average deviation from the mean.

$$
sd = \sqrt{var} = \sqrt{\frac{\sum_{i=1}^n{(\bar{x} - x_i)^2}}{n-1}}
$$

***

#### Challenge

1. Calculate the **variance** of the number of M&Ms in a bag, considering all of the bags in the class. Do this without using the `var` function.

2. Calculate the **standard deviation** of the number of M&Ms in a bag. Do this without using the `sd` function.

***

## Probability and M&Ms

Without looking, you chose one M&M. What colors could you have chosen?

The **set** of brown, yellow, green, red, orange, blue is the **sample space**.

Your selection of a single M&M is called an **event**.

*** 

#### Challenge

What's the probability of getting a "color" M&M in **your** bag?

P(brown), P(yellow), etc.

***

### Putting our data together into a `data.frame`

Here we will collate the classes data into a single data set.

Go to this Google Sheet and add your data:
[https://docs.google.com/spreadsheets/d/1FKOtz2XKOMCBcKX2_g3BTrNsB6CkXwdVbS-Fm3gwaN0/edit?usp=sharing](https://docs.google.com/spreadsheets/d/1FKOtz2XKOMCBcKX2_g3BTrNsB6CkXwdVbS-Fm3gwaN0/edit?usp=sharing)

#### Aside: Reading in your own data

One of the most basic things you will need to do in R is read in your own data set. 
You can read in Excel files, simple text files, and even files from Google Sheets.
But the easiest type of file to read in is a comma separated values (CSV) file. 
You can save an Excel workbook (or Numbers workbook or Google Sheet) as a CSV file by using the "Save as ..." menu item. 

Let's download our M&M data set that we just put together.
First, download the file here. Right click the link and select "Download linked file" - [M&M data](https://github.com/mlammens/ENS-623-Research-Stats/raw/gh-pages/lectures/mm_data.csv) **The file is called `mm_data.csv`.**

Put the file in your `ENS-623-Research-Stats/data` folder. Then we can read in the file.

```{r, eval=FALSE}
read.csv("data/mm_data.csv")

mm_data <- read.csv("data/mm_data.csv")
```

#### Aside: Manipulating a data set with `tidyr` and `dplyr` (enter the `tidyverse`)

[`tidyvers`](https://www.tidyverse.org/) is a set of R packages focused on data science. `ggplot2` is actually one of these packages. 
We're going to use some of the tools in the `tidyverse` to do some basic data manipulation.

There are several very good resources for learning about the tools in the `tidyverse`. 
The most comprehensive is the book linked from the tidyverse website. 

Personally, I rely on this [data wrangling cheetsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).

```{r, eval=FALSE}
install.packages("tidyverse")
```

```{r}
# load tidyverse libraries
library(tidyverse)

# Read in the M&M data set
mm_data <- read.csv("mm_sample.csv")

# look at these data
mm_data

```

Currently, these data are in **wide-format**, which while easy to read, is a bit hard to manipulate.

We want to convert to **long-format**. In this format, each row will correspond to **exactly one observation**.

```{r}
mm_data_long <- tidyr::gather(mm_data, "Person", "MM_Count", 2:4)

# Look at these data in long format
mm_data_long
```


We can use `dplyr` tools to perform various data processing tasks on these data.

**Calculate the total M&Ms in each persons bag**

```{r}
mm_data_long %>%
  group_by(Person) %>%
  summarise(Tot_MM = sum(MM_Count))
```

**Add the sum as another column to our *long* format data set**

```{r}
mm_data_long <-
  mm_data_long %>%
  group_by(Person) %>%
  mutate(Tot_MM = sum(MM_Count))

mm_data_long
```


**Use the above to calculate the prob of getting each color in each bag**

```{r}
mm_data_long$Prob_Color <- round(mm_data_long$MM_Count / mm_data_long$Tot_MM, 2)

mm_data_long
```


***

#### Challenge

What is the mean probability of getting each color across each bag?

Tools to use:

* `data.frames`
* `apply`

***

## Probability of events

* What's the P(green OR blue OR red) in **your** bag? 

P(green) + P(blue) + P(red)

* What is the P(NOT green) in **your** bag?

$$
P(\sim Green) = 1 - P(Green) = P(G)^c
$$


### Sampling with replacement

I draw one M&M from my bag, put it back, then draw another.

* What's the P(green and then blue)?

<!---
P(green) * P(blue) 
--->

* What's the P(green or blue)?

<!---
P(green) + P(blue)
--->

### Sampling without replacement

* What is the probability of getting green, then blue, *without replacing your first draw*?

***

#### Challenge

What is the sample space when drawing two M&Ms?

***

## Draw a random bag of M&Ms using the company stated frequencies/probabilities for each color

***

#### Challenge

How would you create a random bag of M&Ms, assuming that each M&M has an equal probability of being in any given bag?

***

Mars claims that percentages of each color M&Ms are slightly different.

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
mm_probs <- c(.23, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 60, replace = TRUE, prob = mm_probs)
table(new_bag)

```


Because it is possible to draw a bag that is completely missing some of the colors, we need to explicitly check how many of each color is in the new bag, if we want to compare with our original bag.


```{r}
## Count the number of each color
new_bag_counts <- c(sum(new_bag == "blue"),
                    sum(new_bag == "brown"),
                    sum(new_bag == "green"),
                    sum(new_bag == "orange"),
                    sum(new_bag == "red"),
                    sum(new_bag == "yellow"))
new_bag_counts
```

And now to check if my bag is the same as the new bag I can look at a series of logical tests.

```{r, eval=FALSE}
individual_color_compare <- new_bag_counts == my_orig_bag

# The bags match if all of these are true
all(individual_color_compare)
```


