---
title: "Meeting 4 - Working with Data in R"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

> *Note - much of the material in this session is based on or remixed from materials in the [Data Carpentry - Data Analysis and Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/index.html) lessons, maintained by François Michonneau & Auriel Fournier. Copyright (c) Data Carpentry.*

# Today's Goals

* Be able to load a data set into the R environment from a .csv file
* Practice with basic data summary and `data.frame` indexing functions
* Practice working with **factors** and **dates**
* Become familiar with the `tidyverse` "curated" list of packages
* Practice subsetting a data set using functions `select` and `filter`
* Practice calculating summaries of a data set usinig functions `group_by` and `summarize`

***

# Working with data files

## Portal Project data set

We will be working with a data set from the [Portal Project](https://portal.weecology.org/), a long-term ecological research project near Portal, Arizona (southeast AZ) that was started in 1977. 
Data collected as part of this project includes observations of plants, rodents, and ants in a set of experimentally manipulated study plots. 

### Downloading the data

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from [figshare](https://figshare.com/), a website for storing and archiving research products.
Here's a direct link to the figshare page for these data - [https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10](https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10) - but we won't be downloading these data manually.

Inside the `download.file` command, the first entry is a character string with the
source URL ("https://ndownloader.figshare.com/files/2292169"). 
This source URL downloads a CSV file from figshare. The text after the comma
("data/portal_data_joined.csv") is the destination of the file on your local
machine. You'll need to have a folder on your machine called "data" where 
you'll download the file to. 
You should have created this folder in the first week of class.
So this command downloads a file from figshare, names 
it "portal_data_joined.csv" and adds it to a preexisting folder named "data".

```{r, eval=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

```{r, echo=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "../data/portal_data_joined.csv")
```


### Reading the data into R

The file has now been downloaded to the destination you specified, but R has not 
yet loaded the data from the file into memory. 
To do this, we are going to use the `read_csv()` function from the **`tidyverse`** package.
Note that there other ways to read data in, such as the base R `read.csv()`, but we'll use the upgraded `read_csv()` function here.

First, we need to install the **`tidyverse`** package. 
Type `install.packages("tidyverse")` straight into the console. 
It's better to write this in the console than in our script for any package, as there's no 
need to re-install packages every time we run the script. 
Then, to load the package type:

```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

Now we can use the functions from the **`tidyverse`** package. 
Let's use `read_csv()` to read the data into a data frame:

```{r, eval=FALSE,  purl=FALSE}
surveys <- read_csv("data/portal_data_joined.csv")
```

```{r, eval=TRUE, echo=FALSE,  purl=FALSE}
surveys <- read_csv("../data/portal_data_joined.csv")
```


You will see the message `Parsed with column specification`, followed by each
column name and its data type.
When you execute `read_csv` on a data file, it looks through the first 1000 rows 
of each column and guesses its data type. For example, in this data set,
`read_csv()` reads `weight` as `col_double` (a numeric data type), and `species` 
as `col_character`. You have the option to specify the data type for a column
manually by using the `col_types` argument in `read_csv`.

### Examining the `surveys` data frame

Let's use some of the functions we've learned in the past, and a few new ones, to look at the `surveys` data frame.

First, let's open the data frame in a viewer.

```{r, eval = FALSE}
View(surveys)
```

Next, let's use the `summary` function.

```{r}
summary(surveys)
```

How about a few other functions?

***

#### Challenge - looking a data frames

Try using each of the functions below and identifying what they do.

```{r, eval = FALSE}
dim(surveys)
nrow(surveys)
ncol(surveys)
names(surveys)
rownames(surveys)
str(surveys)

```

***

### Metadata

Metadata is information about the data. 
Standards for metadata can be complicated, but at its most basic level, metadata is simply information about each column of a data set. 
Here's an example of metadata for the Portal data set.

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon            | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

### Practice navigating a `data.frame` - REVIEW

We can access elements of a `data.frame` in many ways. 
The challenge below asks you to identify some of these ways.

***

#### Challenge - navigating a `data.frame`

Describe what each of the lines of code is doing and what it is returning to the user.

```{r, eval=FALSE}
surveys[1,1]
surveys[1:200, 1:5]
surveys$species_id
surveys$hindfoot_length[1:10]

```

***

#### Challenge - editing an element of `data.frame`

Execute the code below and describe what each step is doing

```{r}
survey_shrt <- surveys[c(1,8,100,9075,20000), ]
survey_shrt_orig <- survey_shrt
survey_shrt$sex[2] <- "M"
survey_shrt$hindfoot_length[2] <- 320
survey_shrt$weight[1:2] <- mean(survey_shrt$weight, na.rm = TRUE)
```

***

## Factors

When we ran `str(surveys)` above we saw that several of the columns consist of
integers. The columns `genus`, `species`, `sex`, `plot_type`, ... however, are
of the class `character`.
Arguably, these columns contain **categorical data**, that is, they can only take on a limited number of values. 
R has a special class for working with categorical data, called `factor`.

Once created, factors can only contain a pre-defined set of values, known as
**levels**. 
Factors are stored as integers associated with labels and they can be ordered or unordered. While factors look (and often behave) like character vectors, they are actually treated as integer vectors by R. *So you need to be very careful when treating them as strings*.

When importing a data frame with `read_csv()`, the columns that contain text are not automatically coerced (=converted) into the `factor` data type. 
**Note - ** this is different than the behavior of the base R `read.csv()` function.
Once we have loaded the data we can convert columns to factors using the `factor()` function: 

```{r}
surveys$sex <- factor(surveys$sex)
```

We can see that the conversion has worked by using the `summary()` 
function again. This produces a table with the counts for each factor level:

```{r}
summary(surveys$sex)
```

**IMPORTANT -** By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because `f` comes before `m`, even though the first element in this vector is
`"male"`). 
You can see this by using the function `levels()` and you can find the number of levels using `nlevels()`:

```{r}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r}
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more informative than integers because factors are self describing: `"female"`,
`"male"` is more descriptive than `1`, `2`. Which one is "male"?  You wouldn't
be able to tell just from the integer data. Factors, on the other hand, have
this information built in. You can actually see the conversion (and see some of the ways you might run into trouble) by coercing a factor into a number:

```{r}
as.numeric(sex)
```

***

#### Challenge - applying factors to the Portal data

Change the columns `taxa` and `genus` in the `surveys` data frame into a factor. 
Once you have done this, use the `summary` function to determine how many of each `taxa` are in the data set.

***

Here's a more detailed example of how converting factors can go off the rails.

```{r}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
```

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a quick glance at the number of observations represented by each factor level. 
Let's look at the number of males and females captured over the course of the experiment:

```{r}
## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

However, as we saw when we used `summary(surveys$sex)`, there are about 1700 
individuals for which the sex information hasn't been recorded. To show them in
the plot, we can turn the missing values into a factor level with the 
`addNA()` function. We will also have to give the new factor level a label.
We are going to work with a copy of the `sex` column, so we're not modifying the working copy of the data frame:

```{r}
sex <- surveys$sex
levels(sex)
sex <- addNA(sex)
levels(sex)
head(sex)
levels(sex)[3] <- "undetermined"
levels(sex)
head(sex)
```

Now we can plot the data again, using `plot(sex)`.

```{r}
plot(sex)
```



