---
title: "Meeting 8 - R Programming and Data Visualizatoin"
author: "Matthew E. Aiello-Lammens"
date: '2019-03-27'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Introduce some basic programming techniques used in reproducible analysis
* Become familiar with data visualization using R and the `ggplot2` package

***

# R Programming

## User Defined Functions

What is a function?

Why write a function?

**Example**

*Note:* the example below is borrowed from the [Software Carpentry](https://software-carpentry.org/) Introduction to Programming materials.

Convert from Fahrenheit to Kelvin

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

Convert from Kelvin to Celsius

```{r}
kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

#### Challenge

Write a function to convert from Fahrenheit to Celsius.


## `for` loops

How do I do the same thing many times?

### Generic for loop

```
for (variable in collection) {
  do things with variable
}
```

Let's get more specific. Say we took a bunch of measurements of temperature in Fahrenheit, but want to convert them. How might we do it?

#### Make our temperature data set

```{r}
set.seed(8)
temp_data <- runif(n = 20, min = -5, max = 5) + 45
```

#### Challenge

What did we just do?


#### Aside: Resetting the `seed` so the numbers are random again

To reset your seed, you can execute the following command:

```{r}
rm(.Random.seed, envir=globalenv())
```


### Use our `fahr_to_kelvin` function on each element

Iteration 1.

```{r}
for( x in temp_data){
  fahr_to_kelvin(x)
}
```

Iteration 2. By element

```{r}
for( x in temp_data){
  print(fahr_to_kelvin(x))
}
```

Iteration 3. By element location

```{r}
for( x in 1:length(temp_data)){
  print(fahr_to_kelvin(temp_data[x]))
}
```

Iteration 4. By element locatoin with output in a new vector

```{r}
temp_data_kelvin <- vector()
for( x in 1:length(temp_data)){
  temp_data_kelvin[x] <- fahr_to_kelvin(temp_data[x])
}

print(temp_data_kelvin)
```



## Conditionals

We can use conditional statements to **control the flow** of our code, and to "make choices" as it progresses.

### `if` and `else`

The `if` and `else` statements are key to making choices in your code.
Before understanding `if`/`else` statements, we need to review **booleans** - i.e., `TRUE` and `FALSE`

#### Aside: `TRUE` and `FALSE` values

```{r}
TRUE
```

```{r}
20 == 20
```

```{r}
20 > 40
```

A `!` sign can be used as a logical negation

```{r}
!(20 > 40)
```

There are many logical operators to consider.

***

#### Challenge Describe what each fo the following operators are doing.

```{r}
x <- TRUE
y <- FALSE
```

```{r}
x & y
```

```{r}
x | y
```

```{r}
xy <- c(x,y)
```

```{r}
any(xy)
```

```{r}
all(xy)
```

***

Now that we know a bit about booleans, let's get into `if`/`else` statements.

Essentially, an `if` is a conditional that says, "do this thing after the `if` statement, ***if*** the conditional was TRUE."

Here's a simple example (taken from Software Carpentry's lessons)

```{r}
num <- 37
if (num > 100) {
  print("greater than 100")
} else {
  print("not greater than 100")
}
```

#### Challenge 

Re-define `num` so you get the other option.

We don't need an `else` statement for this to work -

```{r}
num <- 37
if (num > 100) {
  print("The number was greater than 100")
}

```

We can also write a "cascade" of `if`/`else` statements

```{r}
if (num > 0) {
  return(1)
} else if (num == 0) {
  return(0)
} else {
  return(-1)
}
```

#### Challenge

Make the above into a function, that takes in any value, and returns whether it is positive, negative, or equal to 0.

***

# Data Visualization and Exploration

The first step to any data analysis is to interrogate your data by calculating some standard statistics and by visualizing it in various ways. 
In future classes we will look more closely at standard statistics.
Today, we'll focus on data visualization.

### Iris dataset

We're going to work with the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html) again.


```{r}
data( iris )
summary(iris)
```


#### Aside: A review of navigating a `data.frame`

#### Challenge

What commands can I use if I want the first 10 elements of both the sepal width and the species columns?


#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

## Visualization using `base` R functions

`base` R is how we refer to the things that come pre-loaded in your R environment. Think of them as the **basics**. We'll talk more about add on **packages** in just a bit.

Let's add a habitat type variable to the iris data set. We'll use this later. 
*Caveat - I made up these habitat type preferences.*

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )

iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )

head( iris_full )
tail( iris_full )
```

### Visualizing the measurements of a single variable

Perhaps the most common way to look at data for a single variable is a histogram.
This essentially is a bar plot, where each bar represents the number of times a value falls within a particular *bin*.

Example: let's look at the distribution of `Petal.Length` values

```{r}
hist(iris_full$Petal.Length)
```

***

#### Challenge - use `?hist` to determine how to change the number of bins used

Hint: bins are sometimes called breaks

```{r}
?hist
hist(iris_full$Petal.Length, breaks = 30)
```

***

### Visualizing relationships between two variables

We would like to plot Sepal.Length versus Petal.Length. We'll first do this using
**base R** functions.

```{r plot1}
plot( x = iris_full$Sepal.Length, y = iris_full$Petal.Length )
```

Note, we could save a little typing by using the `with` function. This 
sets up a temporary environment where all of the variables (columns) 
in my data set are defined individually.

```{r plot1-with}
with( data = iris_full, plot( x = Sepal.Length, y = Petal.Length ) )
```


## Visualization using the `ggplot2` package

Now we're going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved.

#### Aside: Installing and loading packages

First, let's install the `ggplot2` package:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```


OK, let's fire up ggplot.

```{r plot1-ggplot}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) )
```

Let's break down that call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

**THAT SEEMS SO COMPLICATED!**

It's true, for simple plots, ggplot can be much more complicated
than simply using the base functions. But the power of ggplot lies
in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. For example,
let's say I want to create the plot we just made, but have 
each species represented by a different color.

```{r plot2}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) )
```

Let's add more information - how about habitat type as well.

```{r}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species, shape = Habitat), size = 2.5 ) +
  theme_bw()
```


**facets** - a way to separate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`.

```{r plot2-facets}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```

**NOTE**: I moved the `data ...` stuff into the initial `ggplot` call here.

*** 

#### Challenge

1. `ggplot2` has many geometries, allowing us to make lot's of different 
types of plots. Let's make two new plots - one **boxplot** of `Petal.Length`, 
with one boxplot for each species. Use `geom_boxplot` for this.

```{r}
ggplot(data = iris_full, aes(x = Species, y = Petal.Length)) +
  geom_boxplot(aes(fill = Species)) +
  theme_bw()  
```

2. Make a histogram of Petal.width. 

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length)) +
  theme_bw()  
```


Use facets to separate the three species.

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length, fill = Species)) +
  facet_grid( Species ~ . ) +
  theme_bw()  
```

3. Second, let's make density plots of Petal.Width. Use `geom_density` and
`colour` to make different colored density lines for iris in each 
habitat type. 
**Note: the area under the curve is equal to 1.**

```{r}
ggplot() +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()
```

4. Use histogram to plot density instead of counts.

```{r}
ggplot() +
  geom_histogram(data = iris_full, 
                 aes(x = Petal.Length, y = ..density.., fill = Species)) +
  facet_grid( Species ~ . ) +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()  
```


***

### More `ggplot2` resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.

***
