---
title: "Meeting 6 - Probability and Probability Distributions"
author: "Matthew E. Aiello-Lammens"
date: '2020-02-26'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Complete introduction to basic principles of probability
* Practice visualizing probability using Venn diagrams
* Learn concept of conditional probability
* Introduce concepts of probability distributions
* Learn details of a few important probability distributions



***

# Probability through M&Ms (cont'd)


## Mutually Exclusive Outcomes

**Challenge** - what's the probability of getting a green M&M?

$P(green) = 0.16$ -- likewise, we can identify the associated values for each of the other colors.

**Challenge** - what's the probability of your M&M being *both* green **AND** blue?

These events are mutually exclusive.

**Challenge** - What's the probability of getting a green **OR** a blue M&M?

$P(green \space\cup\space blue) = P(green) + P(blue)$

**Challenge** - What's the probability of getting a blue, brown, green, orange, red, or yellow M&M?

### Addition Rule of Mutually Exclusive (Disjoint) Outcomes

Given $A_1$ and $A_2$ are mutually exclusive, then the 

$$
P(A_1\text{ or } A_2 ) = P(A_1) + P(A_2)
$$

Extending this to $k$ mutually exclusive events:

$$
P(A_1) + P(A_2) + ... + P(A_k)
$$

## Non-mutually Exclusive Events

Let's think about rolling a six-sided die for the following examples.

**Example** - 6-sided die

$$
P(<4 \text{ OR odd})
$$

Show probability of each, and areas of overlap.

$$
P(A\text{ or } B ) = P(A) + P(B) - P(A\text{ and }B)
$$


## Independence

> *Two processes are **independent** if knowing the outcome of one provides* no useful information *about the outcome of the other.* - Diez et al. 2016, p. 85

**Challenge** - Is flipping two coins different from flipping one, then the other? 

**Challenge** - You flip 99 heads. What is the probability you will flip a head on the 100th toss?

**Challenge** - For our 6-sided die example above, why aren't $P(<4)$ and $P(\text{odd})$ **NOT** independent events?


### Multiplication Rule for Independent Processes

$$
P(A \text{ and } B) = P(A) \times P(B)
$$


# Visualizing probability - Venn Diagrams

For this, we'll need to install and load the `VennDiagram` package.

```{r, eval=FALSE}
install.packages("VennDiagram")
```

Then to load it into my workspace, and make the functions available:

```{r}
library(VennDiagram)
```


Let's draw a Venn diagram representing the probability of getting a **blue** M&M.
This is written algebraically as $P(Blue)$:

```{r, results='hide'}
## Load VennDiagram package - NB: you will likely have to install this package first
library(VennDiagram)

## Create a Venn diagram
draw.single.venn(area = 0.23, category = "Blue M&M", fill = "blue", alpha = 0.5)
```

Draw a Venn diagram representing the probability of getting a **blue or red** M&M.
This is written algebraically as $P(Blue \cup Red)$:

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.23, area2 = 0.13, cross.area = 0, 
                   category = c("Blue M&M", "Red M&M"),
                   fill = c("blue", "red"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that there is no overlap between these two events. We say that these events are *mutually exclusive*.**

### Adding Peanut and Almond M&Ms to the mix

Let's imagine the scenario that we add an equal amount of Peanut and Almond M&Ms to our bags of M&Ms. 
Now we can ask a question like, *what is the probability of getting a **blue AND peanut** M&M?*
We can write this algebraically as $P(Blue \cap Peanut)$.

Recall that $P(Blue) = 0.23$, and because all three types of M&Ms are represented equally, the $P(Peanut) = 0.33$.

```{r, results='hide'}
grid.newpage()
draw.pairwise.venn(area1 = 0.23, area2 = 0.33, cross.area = 0.0759, 
                   category = c("Blue M&M", "Peanut M&M"),
                   fill = c("blue", "yellow"), alpha = rep(0.5, 2))
```

**IMPORTANT: Note that in this example, the color and type of M&M are *indpendent* from each other.**

#### Challenge

**What is the probability of getting a *blue OR peanut* M&M?** I.e. $P(Blue \cup Peanut)$.

Hint: Note the difference between adding up the area represented in the Venn diagram versus adding the $P(Blue) + P(Peanut)$.

### Conditional probability

#### Challenge

What is the probability of drawing a blue M&M, CONDITIONAL on it being a peanut M&M?

$$
P(Blue|Peanut) = \frac{P(Blue \cap Peanut)}{P(Peanut)}
$$

Calculate this probability.[^1] 

[^1]:The value should be 0.23. **Why?**

***

# Probability distributions 

## The return of the red fox!

**What is the probability that a red fox wins *x* bouts**

Recall [problem set 3](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-3.html) (i.e., the red fox versus arctic fox simulation).
What if we wanted to know what the probability of a red fox winning 60 out of 100 bouts is? 
How might we figure that out?
Let's start by using *a modified portion* of the solution to problem set 3. 
Specifically, let's simulate the number of bouts a red fox wins, given some number of total bouts.

* Given the stated probability of a red fox winning, how many bouts do you think the red fox should win?

```{r, results='hide'}
## Start a `for` loop, to replicate a known number of bouts

## Set the probability that a red fox wins the bout
prob_redfox_win <- 0.6

## Set the number of bouts
bouts <- 100 

## Initiate the number of redfox bout wins
redfox_bout_win <- 0


for(ind in 1:bouts){
  

  ## Draw a random number to simulate the outcome of the bout
  bout_outcome <- runif(1)
  
  ## Use an if statement to determine who won
  if(bout_outcome < prob_redfox_win){ # red fox wins
    ## Print the outcome to the console, so we know who won
    print("red fox wins")
    
    ## Increase the number of redfox_bout_win by 1
    redfox_bout_win <- redfox_bout_win + 1
    
  } else { # arctic fox wins
    ## Print the outcome to the console
    print("arctic fox wins")
    
    ## NOTE: We're only interested in the number of redfox wins, so 
    ## we're not going to bother keeping track of the arctic fox wins.
  }  
} # End the `for` loop

print(paste("Red fox won", redfox_bout_win, "bouts!"))

```

* How does your guess from the point above compare to the results of the simulation?


Next, let's run the simulation multiple times and monitor the different values for the number of red fox bout wins.

```{r, results='hide'}
## Set the probability that a red fox wins the bout
prob_redfox_win <- 0.6

## Set the number of replicates
reps <- 100

## Initiate a vector to store the number of red fox wins for each replicate
redfox_bout_wins_vector <- c()

## Start a for loop to go through the desired number of replicates
for(rep in 1:reps){ 
  
  ## Set the number of bouts
  bouts <- 100 
  
  ## Initiate the number of redfox bout wins
  redfox_bout_win <- 0
  
  ## Start for loop to go through the number of bouts
  for(ind in 1:bouts){
    
    
    ## Draw a random number to simulate the outcome of the bout
    bout_outcome <- runif(1)
    
    ## Use an if statement to determine who won
    if(bout_outcome < prob_redfox_win){ # red fox wins
      ## Print the outcome to the console, so we know who won
      #print("red fox wins")
      
      ## Increase the number of redfox_bout_win by 1
      redfox_bout_win <- redfox_bout_win + 1
      
      } else { # arctic fox wins
        ## Print the outcome to the console
        #print("arctic fox wins")
        
        ## NOTE: We're only interested in the number of redfox wins, so 
        ## we're not going to bother keeping track of the arctic fox wins.
        }  
    } # End the `for` loop
  
  print(paste("Red fox won", redfox_bout_win, "bouts!"))
  
  ## Add the number of redfox_bout_win to redfox_bout_wins_vector
  redfox_bout_wins_vector <- c(redfox_bout_wins_vector, redfox_bout_win)
  
}
```

#### Challenge

How might we visualize the number of red fox bout wins?

```{r}
library(ggplot2)

ggplot() +
  geom_histogram(data = NULL, 
                 aes(x = redfox_bout_wins_vector), 
                 binwidth = 1) +
  theme_bw()
```

## Probability distribution of red fox wins - simulated

How might we visualize the **probability** for the numbers of red fox bout wins?

```{r}
library(ggplot2)

ggplot() +
  geom_histogram(data = NULL, 
                 aes(x = redfox_bout_wins_vector, y = ..density.. ), 
                 binwidth = 1) +
  theme_bw()
```


#### Challenge

Using our last plot, estimate the probability of the red fox winning exactly 60 bouts.

## Probability of 60 wins - analytically

Thinking about probability and counting (see [Meeting 5](http://mlammens.github.io/ENS-623-Research-Stats/lectures/Lecture-5.html)), let's determine the exact probability of the red fox winning exactly 60 bouts.

1. What's the probability of a red fox winning 60 bouts, then losing 40?

$$
P(\text{red fox wins})^{60} * P(\text{red fox loses})^{40}
$$

2. How many different ways can there be 60 wins?

Think about the result of each bout as an *indpendent event*.
How many ways can you line up 100 independent events?

$$
n \cdot (n-1) \cdot (n-2) \cdot \text{...} \cdot 1 = n!
$$

, where $n = 100$.

Next, how many ways can you line up 60 wins and 40 losses?

$$
x! \cdot (n-x)!
$$

, where $x = 60$, or the number of wins.

3. Putting this all together we have the following:

$$
\frac{n!}{x! \cdot (n-x)!} P(\text{red fox wins})^{x} * P(\text{red fox loses})^{n-x}
$$

#### Challange

* Plug in the number using R, and calculate the probability of getting exactly 60 wins.

```{r}
(factorial(100)/(factorial(60)*factorial(100-60))) * 0.6^60 * 0.4^(100-60)
```

* How does this compare to your estimate from above?

## Binomial probability distribution

We just derived the binomial probability distribution.
In fact, the equation above is what's called the **probability density function** for the **Bionomial Distribution**.
The Binomial distribution can be thought of as the 
sum of $n$ Bernoulli distributions (see below), all with the same 
parameterization (i.e., probability of success = $p$).  
This is useful if we want to find the probability of getting a certain number
of successes if your repeat some experiment many times.

$$
{pdf} = P(x | N,h) = \binom{N}{x} h^x (1-h)^{N-x}
$$

*Question:* What is the number of successes, $x$, in $N$ trials,
where the probability of a success is $h$



### Bernoulli Distribution:

The Bernoulli distribution is a very simple distribution that
can be used if we have a single event (or experiment)
that has two possible outcomes, governed by some probability. 
For example, the probability of the outcome of a 
single coin toss with a fair coin 
can be described using a Bernoulli distribution.

${pdf} = P(x) = p$ if $x=1$ and $(1-p)$ if $x=0$, where
$x$ is the event outcome (i.e., heads or tails)

### Binomial distribution in R

Let's look at R's internal Binomial distribution function.

```{r}
?rbinom()

new_data <- rbinom(n = 100, size = 100, prob = 0.6)

ggplot() +
  geom_histogram(data = NULL, 
                 aes(x = new_data, y = ..density.. ), 
                 binwidth = 1) +
  theme_bw()
```

#### Challenge

What happens to the histogram if you increase the observations, i.e. `n`?


***


