---
title: "Meeting 10 - Correlation and Linear Regression"
author: "Matthew E. Aiello-Lammens"
date: '2020-04-15'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Introduce foundations of statistical models
* Become familiar with key concepts of linear modeling
* Introduce concepts of covariation and correlation
* Identify relationship between correlation and linear regression

# Introduction to Linear Models

## Statistical Models

From Logan, p. 151:

> A **statistical model** is an expression that attempts to explain patterns in the observed values of a *response variable* by relating the response variable to a set of *predictor variables* and parameters.

A generic model looks like:

response variable = model + error

### Simple linear model

$$
y = bx + a
$$

where $b$ is the slope and $a$ is the y-intercept.

**[Written Notes](http://mlammens.github.io/ENS-623-Research-Stats/lectures/Lecture-10-Notes.pdf) - pages 1-3**

## Co-variation

Calcualte covariation between sepal length and sepal width, using the `iris` dataset.

```{r}
data(iris)
head(iris)

library(ggplot2)
ggplot(data = iris) +
  geom_point(aes(x = Sepal.Length, y = Sepal.Width, colour = Species)) +
  theme_bw()

covar <- sum((iris$Sepal.Length - mean(iris$Sepal.Length)) * (iris$Sepal.Width - mean(iris$Sepal.Width)))/ (nrow(iris) - 1)

## Confirm
var(x = iris$Sepal.Length, y = iris$Sepal.Width)

```


**[Written Notes](http://mlammens.github.io/ENS-623-Research-Stats/lectures/Lecture-10-Notes.pdf) - Page 4**

## Correlation

Calculate the correlation between sepal length and sepal width, and calculate whether $r$ is significantly different from 0 or not.

```{r}
cor_iris <- covar / (sd(iris$Sepal.Length)*sd(iris$Sepal.Width))

se_cor_iris <- sqrt((1 - cor_iris^2)/(nrow(iris) - 2))

t_cor_iris <- cor_iris/se_cor_iris

## Calculate the probability of getting our t value, or one more extreme
pt(q = t_cor_iris, df = (nrow(iris)-2))*2

## Use `cor` to find correlation value
with(data = iris, cor(x = Sepal.Length, y = Sepal.Width, method = "pearson"))

## Test the correlation
with(data = iris, cor.test(x = Sepal.Length, y = Sepal.Width, method = "pearson"))


```


## Robust correlation

For Pearson's correlation coefficient to be appropriate, the data should have:

1. A linear relationship
2. A bivariate normal distribution

* Spearman's Rank Correlation - calculates correlation on ranks of observed values
* Kendall's Rank Correlation


# Linear regression

Now we assume that any change in **y** is due to a change in **x**.

## Example of linear regression

Effects of starvation and humidity on water loss in the confused flour beetle.
Here, looking at the relationship between humidity and weight loss

```{r}
flr_beetle <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter8/Data/nelson.csv")
flr_beetle
```

Plot these data

```{r}
library(ggplot2)
ggplot() +
  geom_point(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  theme_bw()
```

Run a linear regression

```{r}
flr_beetle_lm <- lm(data = flr_beetle, WEIGHTLOSS ~ HUMIDITY)
summary(flr_beetle_lm)
```

Plot these data, with `lm` fit

```{r}
ggplot(data = flr_beetle, aes( x = HUMIDITY, y = WEIGHTLOSS)) +
  geom_point() +
  stat_smooth(method = "lm") +
  theme_bw()
```


### Linear regression assumptions

#### The big three:

1. **Normality:** The $y_i$ **AND** error values ($\epsilon_i$) are normally distributed. If normality is violated, *p*-values and confidence intervals may be inaccurate and unreliable.
2. **Homogeneity of variance:** The $y_i$ **AND** error values ($\epsilon_i$) have similar levels of variance across all $x_i$ values. 
3. **Independence:** The $y_i$ **AND** error values are independent of *each other*.

#### Other assumptions:

* **Linearity:** The relationship between $x_i$ and $y_i$ is linear (only important when using simple linear regression).
* **Fixed $x$ values:** The $x_i$ values are measured without error. In practice this means the error in the $x_i$ values is much smaller than that in the $y_i$ values.


