---
title: "Meeting 1 - Introduction to R Programming"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    code_folding: hide
    toc: yes
  pdf_document:
    toc: yes
---

# Class goals

* Learn how to develop a research question
* Become proficient using the R statistical programming language
* Learn how to explore data collected during a research project
* Learn and master the standard statistical analyses used in scientific studies
* Understand the connections between statistical analyses and study design
* Become familiar with resources to explore advanced analysis methods

# Class format

* Some lecture - mostly about concepts
* Lots of **live coding** examples - you should follow along, but don't have to
* Challenges - some will be short, some long. Some you work on alone, some with a partner or in groups


# Introduction to R and RStudio

## Why do I have to learn to program in R?

In short, it's "industry standard". 
But the longer answer has to do with replicability (or reproducability) of data analyses.
Documenting your analysis in R, using R scripts or Rmd files allows you to re-run your analysis whenever the need arises, and to share your analysis work flow with others.

## Getting started

### Difference between R and RStudio

In this class, we will be working primarily in RStudio.
So what is the difference between R and RStudio?
R is both a programming language (specifically a statistical analysis programming language) and a software for using that language, while RStudio is an **Integrated Development Environment**, or **IDE** for short.
RStudio offers a number of features, mostly related to visual presentation of information, that make writing and working with R code easier.

### Overall layout

There are four panels in the RStudio interface (though you may only have three open when you first start it), each has valuable information.

* Console / Terminal panel (lower-left)
* Environment / History / Git (upper-right)
* Files / Plots / Packages / Help (lower-right)
* Source / Editor (upper-left)

### File management

Before we do anything in R/RStudio, let's make a new folder on our computers where our class data can reside.
You can use your operating systems file manager (i.e., Finder on Mac and Windows Explorer on Windows) to created a new folder where ever suites you. 

* Setup a `ENS-623-Research-Stats` directory
* Make a `data` directory
* Make a `scripts` directory

### Making an R Project

Go back to RStudio.
Let's make a new R Project associated with your `ENS-623-Research-Stats` directory. 
To make a new project, go to the upper right-hand side of the RStudio interface, where it says **Project: (None)**.
Click the little downward arrow, select "New Project", then select "Existing Directory" from the window that pops up. 
Use the graphical user interface (GUI) to navigate to the `ENS-623-Research-Stats` directory, then select "Create Project".
Once you've created your project, your RStudio session will restart in this project.

There is also now a file in the folder you just made that ends in *.Rproj.
If you double click on this file, RStudio will open, opening this project at the same time.
This is a good way to re-open RStudio for your next session.

#### Setting your working directory

If you open a project using the *.Rproj file, then your R session will automatically set the working directory to your `ENS-623-Research-Stats` folder. If you need to set the working directory manually, here are two ways to do that.

Point-and-click method - Use 'Session' > 'Set Working Directory' > 'Choose Directory'.

Using the R Console:

```{r, eval=FALSE}
setwd("/Users/maiellolammens/Dropbox/Pace/Teaching/ENS-623-Research-Stats/ENS-623-Research-Stats-SP18/")
```


### Getting help

* **Help** panel (lower right corner)
* `help.search`

```{r}
help.search("bar plot")
```

#### Challenge

Use the `help.search` function to search for something in statistics that you think should be in R? Did you find anything?

* I know my function - just give me the details - `?barplot`


## R as calculator

We can use R just like any other calculator.

```{r}
3 + 5
```

There's internal control for order of operations (Please Excuse My Dear Aunt Sally)

```{r}
(3 * 5) + 7
```

```{r}
3 * 5 + 7
```

#### Challenge

Write an example where adding parentheses matters.

### Internal functions

There are a ton of internal functions, and a lot of add-ons.

```{r}
sqrt(4)

abs(-5)

sqrt(-5)
```

## R script file

Use a script file for your work. It's easier to go back to and easy to document.

**Important:** within an R file, you can use the # sign to add comments. Anything written after the # is *not* interpreted when you run the code.

**Important:** we can write commands in our R script file, then run/execute those commands by pressing `CMD` + `RETURN` (Macs) or `CNTR` + `RETURN` (Windows). We can also execute only a section of code by highlighting it and pressing these same key strokes.

#### Challenge

Create a new R script file in your `scripts` directory. Run a few basic commands in the new file.

### Basic file managment in R

```{r}
# What working directory am I in?
getwd()

# Move to a different director?
setwd(".")

# What files are in the current directory?
#dir()
```

#### Things to cover

* Navigating the file path
* Tab completion of file paths
* Tab completion of R commands

#### Challenge

* Try to auto-complete `fil`, what do you find?
* Use the brief help menu that comes up to find a function that starts with `file`, and describe what you think it does.

## Rmd file

Use this to integrate text and R code into the same document.
**I will expect most of your homework assignments as an Rmd file.**

*Practice with Rmd file*

## Variables and objects

There are several basic types of data structures in R.

* **VECTORS**: One-dimensional arrays of numbers, character strings, or logical values (T/F)
* **FACTORS**: One-dimensional arrays of factors (Stop - Let's discuss factors)
* **DATA FRAMES**: Data tables in which the various columns may be of different type
* **MATRICES**: In R, matrices can only be 2-dimensional, anything higher dimension is called an array (see below). Matrix elements are numerical; some functions, like the transpose function t(), only work on matrices
* **ARRAYS**: higher dimensional matrices are called arrays in R
* **LISTS**: lists can contain any type of object as list elements. You can have a list of numbers, a list of matrices, a list of characters, etc., or any combination of the above.

### Functions that are useful for understanding the different types of data structures

```
str()
class()
```

## Practice with variables

Define a variable

```{r}
my_var <- 8
```

And another

```{r}
my_var2 <- 10
```

Work with vars

```{r}
my_var + my_var2
```

Make a new variable

```{r}
my_var_tot <- my_var + my_var2
```

***

#### Challenge

Change the value of my_var2

```{r}
my_var2 <- 3
```

What is the value of `my_var_tot` now?

***

