---
title: "Meeting 4 - Working with Data in R"
author: "Matthew E. Aiello-Lammens"
output:
  html_document:
    toc: yes
---

> *Note - much of the material in this session is based on or remixed from materials in the [Data Carpentry - Data Analysis and Visualization in R for Ecologists](https://datacarpentry.org/R-ecology-lesson/index.html) lessons, maintained by François Michonneau & Auriel Fournier. Copyright (c) Data Carpentry.*

# Today's Goals

* Be able to load a data set into the R environment from a .csv file
* Practice with basic data summary and `data.frame` indexing functions
* Practice working with **factors** and **dates**
* Become familiar with the `tidyverse` "curated" list of packages
* Practice subsetting a data set using functions `select` and `filter`
* Practice calculating summaries of a data set usinig functions `group_by` and `summarize`

***

# Working with data files

## Portal Project data set

We will be working with a data set from the [Portal Project](https://portal.weecology.org/), a long-term ecological research project near Portal, Arizona (southeast AZ) that was started in 1977. 
Data collected as part of this project includes observations of plants, rodents, and ants in a set of experimentally manipulated study plots. 

### Downloading the data

We are going to use the R function `download.file()` to download the CSV file
that contains the survey data from [figshare](https://figshare.com/), a website for storing and archiving research products.
Here's a direct link to the figshare page for these data - [https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10](https://figshare.com/articles/dataset/Portal_Project_Teaching_Database/1314459/10) - but we won't be downloading these data manually.

Inside the `download.file` command, the first entry is a character string with the
source URL ("https://ndownloader.figshare.com/files/2292169"). 
This source URL downloads a CSV file from figshare. The text after the comma
("data/portal_data_joined.csv") is the destination of the file on your local
machine. You'll need to have a folder on your machine called "data" where 
you'll download the file to. 
You should have created this folder in the first week of class.
So this command downloads a file from figshare, names 
it "portal_data_joined.csv" and adds it to a preexisting folder named "data".

```{r, eval=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "data/portal_data_joined.csv")
```

```{r, echo=FALSE, purl=TRUE}
download.file(url = "https://ndownloader.figshare.com/files/2292169",
              destfile = "../data/portal_data_joined.csv")
```


### Reading the data into R

The file has now been downloaded to the destination you specified, but R has not 
yet loaded the data from the file into memory. 
To do this, we are going to use the `read_csv()` function from the **`tidyverse`** package.
Note that there other ways to read data in, such as the base R `read.csv()`, but we'll use the upgraded `read_csv()` function here.

First, we need to install the **`tidyverse`** package. 
Type `install.packages("tidyverse")` straight into the console. 
It's better to write this in the console than in our script for any package, as there's no 
need to re-install packages every time we run the script. 
Then, to load the package type:

```{r, message = FALSE, purl = FALSE}
## load the tidyverse packages, incl. dplyr
library(tidyverse)
```

Now we can use the functions from the **`tidyverse`** package. 
Let's use `read_csv()` to read the data into a data frame:

```{r, eval=FALSE,  purl=FALSE}
surveys <- read_csv("data/portal_data_joined.csv")
```

```{r, eval=TRUE, echo=FALSE,  purl=FALSE}
surveys <- read_csv("../data/portal_data_joined.csv")
```


You will see the message `Parsed with column specification`, followed by each
column name and its data type.
When you execute `read_csv` on a data file, it looks through the first 1000 rows 
of each column and guesses its data type. For example, in this data set,
`read_csv()` reads `weight` as `col_double` (a numeric data type), and `species` 
as `col_character`. You have the option to specify the data type for a column
manually by using the `col_types` argument in `read_csv`.

### Examining the `surveys` data frame

Let's use some of the functions we've learned in the past, and a few new ones, to look at the `surveys` data frame.

First, let's open the data frame in a viewer.

```{r, eval = FALSE}
View(surveys)
```

Next, let's use the `summary` function.

```{r}
summary(surveys)
```

How about a few other functions?

***

#### Challenge - looking a data frames

Try using each of the functions below and identifying what they do.

```{r, eval = FALSE}
dim(surveys)
nrow(surveys)
ncol(surveys)
names(surveys)
rownames(surveys)
str(surveys)

```

***

### Metadata

Metadata is information about the data. 
Standards for metadata can be complicated, but at its most basic level, metadata is simply information about each column of a data set. 
Here's an example of metadata for the Portal data set.

| Column           | Description                        |
|------------------|------------------------------------|
| record\_id       | Unique id for the observation      |
| month            | month of observation               |
| day              | day of observation                 |
| year             | year of observation                |
| plot\_id         | ID of a particular plot            |
| species\_id      | 2-letter code                      |
| sex              | sex of animal ("M", "F")           |
| hindfoot\_length | length of the hindfoot in mm       |
| weight           | weight of the animal in grams      |
| genus            | genus of animal                    |
| species          | species of animal                  |
| taxon            | e.g. Rodent, Reptile, Bird, Rabbit |
| plot\_type       | type of plot                       |

### Practice navigating a `data.frame` - REVIEW

We can access elements of a `data.frame` in many ways. 
The challenge below asks you to identify some of these ways.

***

#### Challenge - navigating a `data.frame`

Describe what each of the lines of code is doing and what it is returning to the user.

```{r, eval=FALSE}
surveys[1,1]
surveys[1:200, 1:5]
surveys$species_id
surveys$hindfoot_length[1:10]

```

***

#### Challenge - editing an element of `data.frame`

Execute the code below and describe what each step is doing

```{r}
survey_shrt <- surveys[c(1,8,100,9075,20000), ]
survey_shrt_orig <- survey_shrt
survey_shrt$sex[2] <- "M"
survey_shrt$hindfoot_length[2] <- 320
survey_shrt$weight[1:2] <- mean(survey_shrt$weight, na.rm = TRUE)
```

***

## Factors

When we ran `str(surveys)` above we saw that several of the columns consist of
integers. The columns `genus`, `species`, `sex`, `plot_type`, ... however, are
of the class `character`.
Arguably, these columns contain **categorical data**, that is, they can only take on a limited number of values. 
R has a special class for working with categorical data, called `factor`.

Once created, factors can only contain a pre-defined set of values, known as
**levels**. 
Factors are stored as integers associated with labels and they can be ordered or unordered. While factors look (and often behave) like character vectors, they are actually treated as integer vectors by R. *So you need to be very careful when treating them as strings*.

When importing a data frame with `read_csv()`, the columns that contain text are not automatically coerced (=converted) into the `factor` data type. 
**Note - ** this is different than the behavior of the base R `read.csv()` function.
Once we have loaded the data we can convert columns to factors using the `factor()` function: 

```{r}
surveys$sex <- factor(surveys$sex)
```

We can see that the conversion has worked by using the `summary()` 
function again. This produces a table with the counts for each factor level:

```{r}
summary(surveys$sex)
```

**IMPORTANT -** By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r}
sex <- factor(c("male", "female", "female", "male"))
```

R will assign `1` to the level `"female"` and `2` to the level `"male"` (because `f` comes before `m`, even though the first element in this vector is
`"male"`). 
You can see this by using the function `levels()` and you can find the number of levels using `nlevels()`:

```{r}
levels(sex)
nlevels(sex)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `sex` vector would be:

```{r}
sex # current order
sex <- factor(sex, levels = c("male", "female"))
sex # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2, 3), but are more informative than integers because factors are self describing: `"female"`,
`"male"` is more descriptive than `1`, `2`. Which one is "male"?  You wouldn't
be able to tell just from the integer data. Factors, on the other hand, have
this information built in. You can actually see the conversion (and see some of the ways you might run into trouble) by coercing a factor into a number:

```{r}
as.numeric(sex)
```

***

#### Challenge - applying factors to the Portal data

Change the columns `taxa` and `genus` in the `surveys` data frame into a factor. 
Once you have done this, use the `summary` function to determine how many of each `taxa` are in the data set.

***

Here's a more detailed example of how converting factors can go off the rails.

```{r}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)               # Wrong! And there is no warning...
as.numeric(as.character(year_fct)) # Works...
as.numeric(levels(year_fct))[year_fct]    # The recommended way.
```

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a quick glance at the number of observations represented by each factor level. 
Let's look at the number of males and females captured over the course of the experiment:

```{r}
## bar plot of the number of females and males captured during the experiment:
plot(surveys$sex)
```

However, as we saw when we used `summary(surveys$sex)`, there are about 1700 
individuals for which the sex information hasn't been recorded. To show them in
the plot, we can turn the missing values into a factor level with the 
`addNA()` function. We will also have to give the new factor level a label.
We are going to work with a copy of the `sex` column, so we're not modifying the working copy of the data frame:

```{r}
sex <- surveys$sex
levels(sex)
sex <- addNA(sex)
levels(sex)
head(sex)
levels(sex)[3] <- "undetermined"
levels(sex)
head(sex)
```

Now we can plot the data again, using `plot(sex)`.

```{r}
plot(sex)
```

## Formatting dates

One of the most common issues that new (and experienced!) R users have is
converting date and time information into a variable that is appropriate and
usable during analyses. As a reminder from earlier in this lesson, the best
practice for dealing with date data is to ensure that each component of your
date is stored as a separate variable. Using `str()`, We can confirm that our
data frame has a separate column for day, month, and year, and that each contains integer values.

```{r}
str(surveys)
```

We are going to use the `ymd()` function from the package **`lubridate`** (which belongs to the **`tidyverse`**; learn more [here](https://www.tidyverse.org/)). **`lubridate`** gets installed as part as the **`tidyverse`** installation. When you load  the **`tidyverse`** (`library(tidyverse)`), the core packages (the packages used in most data analyses) get loaded. **`lubridate`** however does not belong to the core tidyverse, so you have to load it explicitly with `library(lubridate)`

Start by loading the required package:

```{r}
library(lubridate)
```

`ymd()` takes a vector representing year, month, and day, and converts it to a
`Date` vector. `Date` is a class of data recognized by R as being a date and can
be manipulated as such. The argument that the function requires is flexible,
but, as a best practice, is a character vector formatted as "YYYY-MM-DD".


Let's create a date object and inspect the structure:

```{r}
my_date <- ymd("2015-01-01")
str(my_date)
```

Now let's paste the year, month, and day separately - we get the same result:

```{r}
# sep indicates the character to use to separate each component
my_date <- ymd(paste("2015", "1", "1", sep = "-")) 
str(my_date)
```

Now we apply this function to the surveys dataset. Create a character vector from the `year`, `month`, and `day` columns of
`surveys` using `paste()`:

```{r, eval = FALSE}
paste(surveys$year, surveys$month, surveys$day, sep = "-")
```

This character vector can be used as the argument for `ymd()`:

```{r, eval = FALSE}
ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
```

There is a warning telling us that some dates could not be parsed (understood)
by the `ymd()` function. For these dates, the function has returned `NA`, which
means they are treated as missing values.
We will deal with this problem later, but first we add the resulting `Date`
vector to the `surveys` data frame as a new column called `date`:

```{r}
surveys$date <- ymd(paste(surveys$year, surveys$month, surveys$day, sep = "-"))
str(surveys) # notice the new column, with 'date' as the class
```

Let's make sure everything worked correctly. Again, one way to inspect the new column is to use `summary()`:

```{r}
summary(surveys$date)
```

Let's investigate why some dates could not be parsed.

We can use the functions we saw previously to deal with missing data to identify
the rows in our data frame that are failing. If we combine them with what we learned about subsetting data frames earlier, we can extract the columns "year, "month", "day" from the records that have `NA` in our new column `date`. We will also use `head()` so we don't clutter the output:

```{r}
missing_dates <- surveys[is.na(surveys$date), c("year", "month", "day")]
head(missing_dates)
```

***

#### Challenge - problems with dates

Why did these dates fail to parse? If you had to use these data for your
analyses, how would you deal with this situation?

***

# Data manipulation using **`dplyr`** and **`tidyr`**

Bracket subsetting is handy, but it can be cumbersome and difficult to read,
especially for complicated operations. Enter **`dplyr`**. **`dplyr`** is a package for
making tabular data manipulation easier. It pairs nicely with **`tidyr`** which enables you to swiftly convert between different data formats for plotting and analysis.

The **`tidyverse`** package is an
"umbrella-package" that installs **`tidyr`**, **`dplyr`**, and several other packages useful for data analysis, such as  **`ggplot2`**, **`tibble`**, etc.

The **`tidyverse`** package tries to address 3 common issues that arise when
doing data analysis with some of the functions that come with R:

1. The results from a base R function sometimes depend on the type of data.
2. Using R expressions in a non standard way, which can be confusing for new
   learners.
3. Hidden arguments, having default operations that new learners are not aware
   of.

## What are **`dplyr`** and **`tidyr`**?

The package **`dplyr`** provides easy tools for the most common data manipulation
tasks. It is built to work directly with data frames, with many common tasks
optimized by being written in a compiled language (C++). An additional feature is the
ability to work directly with data stored in an external database. The benefits of
doing this are that the data can be managed natively in a relational database,
queries can be conducted on that database, and only the results of the query are returned.

**Advanced data science concept ** - The above addresses a common problem with R in that all operations are conducted
in-memory and thus the amount of data you can work with is limited by available
memory. The database connections essentially remove that limitation in that you
can connect to a database of many hundreds of GB, conduct queries on it directly, and pull back into R only what you need for analysis.

The package **`tidyr`** addresses the common problem of wanting to reshape your data for plotting and use by different R functions. 
Sometimes we want data sets where we have one row per measurement. 
Sometimes we want a data frame where each measurement type has its own column, and rows are instead more aggregated groups
(e.g., a time period, an experimental unit like a plot or a batch number).
Moving back and forth between these formats is non-trivial, and **`tidyr`** gives you tools for this and more sophisticated  data manipulation.

To learn more about **`dplyr`** and **`tidyr`** after the workshop, you may want to check out this
[handy data transformation with **`dplyr`** cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
and this [one about **`tidyr`**](https://github.com/rstudio/cheatsheets/raw/master/data-import.pdf).

I also heavily rely on this [data wrangling cheatsheet](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf) which combines main functions of both `dplyr` and `tidyr`.

Next, we're going to learn some of the most common **`dplyr`** functions:

- `select()`: subset columns
- `filter()`: subset rows on conditions
- `mutate()`: create new columns by using information from other columns
- `group_by()` and `summarize()`: create summary statistics on grouped data
- `arrange()`: sort results
- `count()`: count discrete values

## Selecting columns and filtering rows

To select columns of a data frame, use `select()`. The first argument
to this function is the data frame (`surveys`), and the subsequent
arguments are the columns to keep.

```{r}
select(surveys, plot_id, species_id, weight)
```

To select all columns *except* certain ones, put a "-" in front of
the variable to exclude it.

```{r}
select(surveys, -record_id, -species_id)
```

This will select all the variables in `surveys` except `record_id`
and `species_id`.

To choose rows based on a specific criterion, use `filter()`:

```{r}
filter(surveys, year == 1995)
```


## Pipes

What if you want to select and filter at the same time? There are three
ways to do this: use intermediate steps, nested functions, or pipes.

With intermediate steps, you create a temporary data frame and use
that as input to the next function, like this:

```{r}
surveys2 <- filter(surveys, weight < 5)
surveys_sml <- select(surveys2, species_id, sex, weight)
```

This is readable, but can clutter up your workspace with lots of objects that you have to name individually. With multiple steps, that can be hard to keep track of.

You can also nest functions (i.e. one function inside of another), like this:

```{r}
surveys_sml <- select(filter(surveys, weight < 5), species_id, sex, weight)
```

This is handy, but can be difficult to read if too many functions are nested, as
R evaluates the expression from the inside out (in this case, filtering, then selecting).

The last option, *pipes*, are a recent addition to R. Pipes let you take
the output of one function and send it directly to the next, which is useful
when you need to do many things to the same dataset.  Pipes in R look like
`%>%` and are made available via the **`magrittr`** package, installed automatically
with **`dplyr`**. 

```{r}
surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
```

In the above code, we use the pipe to send the `surveys` dataset first through
`filter()` to keep rows where `weight` is less than 5, then through `select()`
to keep only the `species_id`, `sex`, and `weight` columns. Since `%>%` takes
the object on its left and passes it as the first argument to the function on
its right, we don't need to explicitly include the data frame as an argument
to the `filter()` and `select()` functions any more.

Some may find it helpful to read the pipe like the word "then". For instance,
in the above example, we took the data frame `surveys`, *then* we `filter`ed
for rows with `weight < 5`, *then* we `select`ed columns `species_id`, `sex`,
and `weight`. The **`dplyr`** functions by themselves are somewhat simple,
but by combining them into linear workflows with the pipe, we can accomplish
more complex manipulations of data frames.

If we want to create a new object with this smaller version of the data, we
can assign it a new name:

```{r}
surveys_sml <- surveys %>%
  filter(weight < 5) %>%
  select(species_id, sex, weight)
surveys_sml
```

Note that the final data frame is the leftmost part of this expression.

***

#### Challenge - practicing with pipes

Using pipes, subset the `surveys` data to include animals collected before 1995 and retain only the columns `year`, `sex`, and `weight`.

***

### Mutate

Frequently you'll want to create new columns based on the values in existing
columns, for example to do unit conversions, or to find the ratio of values in two
columns. For this we'll use `mutate()`.

To create a new column of weight in kg:

```{r}
surveys %>%
  mutate(weight_kg = weight / 1000)
```

You can also create a second new column based on the first new column within the same call of `mutate()`:

```{r}
surveys %>%
  mutate(weight_kg = weight / 1000,
         weight_lb = weight_kg * 2.2)
```

If this runs off your screen and you just want to see the first few rows, you
can use a pipe to view the `head()` of the data. (Pipes work with non-**`dplyr`**
functions, too, as long as the **`dplyr`** or `magrittr` package is loaded).

```{r}
surveys %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

The first few rows of the output are full of `NA`s, so if we wanted to remove
those we could insert a `filter()` in the chain:

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  mutate(weight_kg = weight / 1000) %>%
  head()
```

`is.na()` is a function that determines whether something is an `NA`. The `!`
symbol negates the result, so we're asking for every row where weight *is not* an `NA`.

### Split-apply-combine data analysis and the `summarize()` function

Many data analysis tasks can be approached using the *split-apply-combine*
paradigm: split the data into groups, apply some analysis to each group, and
then combine the results. **`dplyr`** makes this very easy through the use of the
`group_by()` function.


#### The `summarize()` function (or `summarise`)

`group_by()` is often used together with `summarize()`, which collapses each
group into a single-row summary of that group.
**Important note - ** the functions `summarize()` and `summarise()` are equivelant. Just two different spellings for the same thing ¯\\\_(ツ)\_/¯.

`group_by()` takes as arguments the column names that contain the **categorical** variables for which you want
to calculate the summary statistics. So to compute the mean `weight` by sex:

```{r}
surveys %>%
  group_by(sex) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```

You can also group by multiple columns:

```{r}
surveys %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight, na.rm = TRUE))
```


We can see that the `sex` column contains 
`NA` values because some animals had escaped before their sex and body weights 
could be determined. The resulting `mean_weight` column does not contain `NA` but 
`NaN` (which refers to "Not a Number") because `mean()` was called on a vector of 
`NA` values while at the same time setting `na.rm = TRUE`. To avoid this, we can 
remove the missing values for weight before we attempt to calculate the summary statistics on weight. Because the missing values are removed first, we can omit `na.rm = TRUE` when computing the mean:

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight))
```


Here are a few more examples of using `group_by` and `summarise`

It is sometimes useful to rearrange the result of a query to inspect the values. For instance, we can sort on `min_weight` to put the lighter species first:

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(min_weight)
```

To sort in descending order, we need to add the `desc()` function. If we want to sort the results by decreasing order of mean weight:

```{r}
surveys %>%
  filter(!is.na(weight)) %>%
  group_by(sex, species_id) %>%
  summarize(mean_weight = mean(weight),
            min_weight = min(weight)) %>%
  arrange(desc(mean_weight))
```

#### Counting

When working with data, we often want to know the number of observations found
for each factor or combination of factors. For this task, **`dplyr`** provides
`count()`. For example, if we wanted to count the number of rows of data for
each sex, we would do:

```{r}
surveys %>%
    count(sex) 
```

The `count()` function is shorthand for something we've already seen: grouping by a variable, and summarizing it by counting the number of observations in that group. In other words, `surveys %>% count()` is equivalent to:  

```{r}
surveys %>%
    group_by(sex) %>%
    summarise(count = n())
```

For convenience, `count()` provides the `sort` argument:  

```{r}
surveys %>%
    count(sex, sort = TRUE) 
```

Previous example shows the use of `count()` to count the number of rows/observations 
for *one* factor (i.e., `sex`). 
If we wanted to count *combination of factors*, such as `sex` and `species`, 
we would specify the first and the second factor as the arguments of `count()`:

```{r}
surveys %>%
  count(sex, species) 
```

With the above code, we can proceed with `arrange()` to sort the table 
according to a number of criteria so that we have a better comparison. 
For instance, we might want to arrange the table above in (i) an alphabetical order of 
the levels of the species and (ii) in descending order of the count:

```{r}
surveys %>%
  count(sex, species) %>%
  arrange(species, desc(n))
```

From the table above, we may learn that, for instance, there are 75 observations of 
the *albigula* species that are not specified for its sex (i.e. `NA`).

## Reshaping with gather and spread

In `surveys`, the rows of `surveys` contain the values of variables associated
with each record (the unit), values such as the weight or sex of each animal 
associated with each record. What if instead of comparing records, we 
wanted to compare the different mean weight of each genus between plots? (Ignoring `plot_type` for simplicity).

We'd need to create a new table where each row (the unit) is comprised of values of variables associated with each plot. In practical terms this means the values
in `genus` would become the names of column variables and the cells would contain the values of the mean weight observed on each plot.

Having created a new table, it is therefore straightforward to explore the 
relationship between the weight of different genera within, and between, the
plots. The key point here is that we are still following a tidy data structure,
but we have **reshaped** the data according to the observations of interest:
average genus weight per plot instead of recordings per date.

The opposite transformation would be to transform column names into values of
a variable.

We can do both these of transformations with two `tidyr` functions, `spread()`
and `gather()`.

#### Spreading

`spread()` takes three principal arguments:

1. the data 
2. the *key* column variable whose values will become new column names.  
3. the *value* column variable whose values will fill the new column variables.

Further arguments include `fill` which, if set, fills in missing values with 
the value provided.

Let's use `spread()` to transform surveys to find the mean weight of each 
genus in each plot over the entire survey period. We use `filter()`,
`group_by()` and `summarise()` to filter our observations and variables of 
interest, and create a new variable for the `mean_weight`.

```{r}
surveys_gw <- surveys %>%
  filter(!is.na(weight)) %>%
  group_by(plot_id, genus) %>%
  summarize(mean_weight = mean(weight))
str(surveys_gw)
```

This yields `surveys_gw` where the observations for each plot are spread across
multiple rows, 196 observations of 3 variables. 
Using `spread()` to key on `genus` with values from `mean_weight` this becomes
24 observations of 11 variables, one row for each plot.

```{r}
surveys_spread <- surveys_gw %>%
  spread(key = genus, value = mean_weight)
str(surveys_spread)
```


We could now plot comparisons between the weight of genera (one is called a genus, multiple are called genera) in different plots, 
although we may wish to fill in the missing values first.

```{r}
surveys_gw %>%
  spread(genus, mean_weight, fill = 0) %>%
  head()
```

#### Gathering

The opposing situation could occur if we had been provided with data in the
form of `surveys_spread`, where the genus names are column names, but we 
wish to treat them as values of a genus variable instead.

In this situation we are gathering the column names and turning them into a
pair of new variables. One variable represents the column names as values, and
the other variable contains the values previously associated with the column names.

`gather()` takes four principal arguments:

1. the data
2. the *key* column variable we wish to create from column names.
3. the *values* column variable we wish to create and fill with values 
associated with the key.
4. the names of the columns we use to fill the key variable (or to drop).

To recreate `surveys_gw` from `surveys_spread` we would create a key called
`genus` and value called `mean_weight` and use all columns except `plot_id` for
the key variable. Here we exclude `plot_id` from being `gather()`ed.

```{r}
surveys_gather <- surveys_spread %>%
  gather(key = "genus", value = "mean_weight", -plot_id)
str(surveys_gather)
```


Note that now the `NA` genera are included in the re-gathered format. Spreading
and then gathering can be a useful way to balance out a dataset so every
replicate has the same composition.

We could also have used a specification for what columns to include. This can be
useful if you have a large number of identifying columns, and it's easier to
specify what to gather than what to leave alone. And if the columns are directly
adjacent, we don't even need to list them all out - just use the `:` operator!

```{r}
surveys_spread %>%
  gather(key = "genus", value = "mean_weight", Baiomys:Spermophilus) %>%
  head()
```

# Exporting data

Now that you have learned how to use **`dplyr`** to extract information from
or summarize your raw data, you may want to export these new data sets to share
them with your collaborators or for archival.

Similar to the `read_csv()` function used for reading CSV files into R, there is
a `write_csv()` function that generates CSV files from data frames.

Before using `write_csv()`, we are going to create a new folder, `data`,
in our working directory that will store this generated dataset. We don't want
to write generated datasets in the same directory as our raw data. It's good
practice to keep them separate. The `data_raw` folder should only contain the raw,
unaltered data, and should be left alone to make sure we don't delete or modify
it. In contrast, our script will generate the contents of the `data`
directory, so even if the files it contains are deleted, we can always
re-generate them.

In preparation for our next lesson on plotting, we are going to prepare a
cleaned up version of the data set that doesn't include any missing data.

Let's start by removing observations of animals for which `weight` and `hindfoot_length` are missing, or the `sex` has not been determined:


```{r}
surveys_complete <- surveys %>%
  filter(!is.na(weight),           # remove missing weight
         !is.na(hindfoot_length),  # remove missing hindfoot_length
         !is.na(sex))                # remove missing sex
```

Now that our data set is ready, we can save it as a CSV file in our `data`
folder.

```{r, eval=FALSE}
write_csv(surveys_complete, file = "data/surveys_complete.csv")
```

