---
title: "Meeting 8 - R Programming and Data Visualizatoin"
author: "Matthew E. Aiello-Lammens"
date: '2019-03-27'
output:
  html_document:
    code_folding: hide
    toc: yes
---

# Today's Goals

* Introduce some basic programming techniques used in reproducible analysis
* Become familiar with data visualization using R and the `ggplot2` package

***

***

### Make a vector

Combining values into a vector

```{r}
# Vector of variables
my_vect <- c(my_var, my_var2)

# Numeric vector
v1 <- c(10, 2, 8, 7, 11, 15)

# Char vector
pets <- c("cat", "dog", "rabbit", "pig")
```

Making a vector of numbers in sequence

```{r}
v2 <- 1:10
v3 <- seq(from = 1, to = 10)
```

#### Challenge

1. Look up the help for the `seq` function, and use this to make a vector from 1 to 100, by steps of 5.
2. Come up with a way that you would use the `length.out` argument.

## Exploring variable elements

You can get specific elements from vectors and other data structures

* Introduction to the square brackets `[]`

```{r}
pets <- c("cat", "dog", "rabbit", "pig", "snake")
pets[1]
```

* Getting a number of elements, in sequence

```{r}
pets[3:4]
```

* Getting a number of elements, not in sequence

```{r}
pets[c(1,4)]
```


## Working with matrices

**Review** - Why might we want 2D data?

Let's make a matrix

#### Challenge

With the people next to you, break down this function, and describe each argument.
What is the final product?

```{r}
my_mat <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

What does it mean to fill `byrow`?

```{r}
matrix(data = 1:9, nrow = 3, byrow = TRUE)
```

Versus

```{r}
matrix(data = 1:9, nrow = 3, byrow = FALSE)
```

#### Challenge

What is the default value for `byrow`?

### Indexing matrices

Indexing happens by **row**, **column** notation.

```{r}
my_mat <- matrix(data = 1:50, nrow = 10, byrow = TRUE)

my_mat[1,1]
my_mat[1,2]
my_mat[2,1]

my_mat[1:4, 1:3]

my_mat[c(1,3,5), ]
my_mat[ ,c(1,3,5)]
```

## Combinining internal functions with matrices

Make a "random" matrix (that isn't random in this case because of the `set.seed` function)

```{r}
set.seed(1)
mat1 <- matrix(data = runif(50), nrow = 10, byrow = TRUE)
```

Calculate the mean of all of the data

```{r}
mean(mat1)
```

Calculate the standard deviation of all of the data

```{r}
sd(mat1)
```

Calculate row means and column means

```{r}
rowMeans(mat1)
colMeans(mat1)
```

Introduce the `apply` function

```{r}
apply(mat1, MARGIN = 1, mean)
```

## Data frames

### Iris dataset

We're going to work with a dataset that comes built in to R, commonly 
called the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html).
It is also sometimes called Fisher's Iris dataset (but should more appropriately
be called Anderson's Iris dataset). Because it comes pre-packaged with R, we 
can load it into our environment using the `data` function.

```{r}
data(iris)
```

Let's have a look at this dataset

```{r}
head(iris)
tail(iris)
```

The dataset contains measurements of four characteristics of three
different species of Iris (plants!). 

#### `summary` function

Let's begin by using the `summary` function to examine this dataset.
`summary` returns many of the standard statistics.
When doing data exploration, a few things you want to look at are:

* How to the mean and median values within a variable compare?
* Do the min and max values suggest there are outliers?
* Which variables (i.e., columns) are quantitative (numeric) versus categorical (factors or characters)

```{r}
summary(iris)
```

#### Aside: Differences between characters and factors

Factors 'appear' to be similar to characters, but are in fact coded numerically in R. 
Think of **factors** like **categories**. 
Here's a quick example that demonstrates the difference in these two variable types that shows up when using `summary`.

```{r}
## Make a new iris dataset
iris_new <- iris

## Create a new column that treats species as a character, rather than a factor
iris_new$species_char <- as.character(iris_new$Species)

## Run summary command
summary(iris_new)
```


#### Aside: A (very) brief introduction to navigating a `data.frame`

We will be very brief here. 
I recommend checking out 
[this Data Carpentry lesson](http://www.datacarpentry.org/R-ecology/03-data-frames.html) for more information.

* Looking at specific `data.frame` elements. Use the *row* and *column* notation.

Here is the 5th row, 3rd column (Petal.Length).
**Note: We are using square brackets to index the `data.frame` and we *always* use row, column notation.**

```{r}
iris[5, 3]
```

* Looking at an entire column.

Here are two ways to get the `Petal.Length` column.

First, **note: we leave the row part blank, but still add the comma.**

```{r}
iris[ ,3]
```

Second, **use only the variable (column) name. Note the use of the `$` operator**

```{r}
iris$Petal.Length
```

* Looking at specific column entry

This is another way to look at the 5th entry in the `Petal.Length` column.

```{r}
iris$Petal.Length[5]
```

* Looking at all entries for a given row.

Here's all the entries for the 5th row.
**Note: here we leave the column part blank, but still add the comma.**

```{r}
iris[5, ]
```

* Looking at a set of rows and/or columns.

Here's all the entries in the 5th through 10th rows, 1st through 3rd columns.
**Note: we use the `:` operator to look at a range of value.**

```{r}
iris[5:10, 1:3]
```

* For `data.frame`s, if you do not use row, column notation, you will get only the columns back.

```{r}
head(iris[2:3])
```

***

#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

## Reading in your own data

One of the most basic things you will need to do in R is read in your own data set. 
You can read in Excel files, simple text files, and even files from Google Sheets.
But the easiest type of file to read in is a comma seperated values (CSV) file. 
You can save an Excel workbook (or Numbers workbook or Google Sheet) as a CSV file by using the "Save as ..." menu item. 

For today, we're going to work with a data set related to gastropod fecundity.
We'll link to the file directly using a Dropbox link, but it's possible to download the file here - [gastropod data](https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter6/Data/ward.csv) **The download will begin when you press the hyperlink! The file is called `ward.csv`.**

### Differences in fecundity of intertidal gastropods in two different intertidal zones (Example 6A in Logan, Box 3.1 in Q&K)

Ward and Quinn (1988) investigated the differences in fecundity of *Lepsiella vinosa* in two different intertidal zones (mussel zone and littorinid zone). 

Get the data and have a quick look

```{r}
gastro <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter6/Data/ward.csv")
summary(gastro)
```


***

# R Programming

## User Defined Functions

What is a function?

Why write a function?

**Example**

*Note:* the example below is borrowed from the [Software Carpentry](https://software-carpentry.org/) Introduction to Programming materials.

Convert from Fahrenheit to Kelvin

```{r}
fahr_to_kelvin <- function(temp) {
  kelvin <- ((temp - 32) * (5 / 9)) + 273.15
  return(kelvin)
}
```

Convert from Kelvin to Celsius

```{r}
kelvin_to_celsius <- function(temp) {
  celsius <- temp - 273.15
  return(celsius)
}
```

#### Challenge

Write a function to convert from Fahrenheit to Celsius.


## `for` loops

How do I do the same thing many times?

### Generic for loop

```
for (variable in collection) {
  do things with variable
}
```

Let's get more specific. Say we took a bunch of measurements of temperature in Fahrenheit, but want to convert them. How might we do it?

#### Make our temperature data set

```{r}
set.seed(8)
temp_data <- runif(n = 20, min = -5, max = 5) + 45
```

#### Challenge

What did we just do?


#### Aside: Resetting the `seed` so the numbers are random again

To reset your seed, you can execute the following command:

```{r}
rm(.Random.seed, envir=globalenv())
```


### Use our `fahr_to_kelvin` function on each element

Iteration 1.

```{r}
for( x in temp_data){
  fahr_to_kelvin(x)
}
```

Iteration 2. By element

```{r}
for( x in temp_data){
  print(fahr_to_kelvin(x))
}
```

Iteration 3. By element location

```{r}
for( x in 1:length(temp_data)){
  print(fahr_to_kelvin(temp_data[x]))
}
```

Iteration 4. By element locatoin with output in a new vector

```{r}
temp_data_kelvin <- vector()
for( x in 1:length(temp_data)){
  temp_data_kelvin[x] <- fahr_to_kelvin(temp_data[x])
}

print(temp_data_kelvin)
```



## Conditionals

We can use conditional statements to **control the flow** of our code, and to "make choices" as it progresses.

### `if` and `else`

The `if` and `else` statements are key to making choices in your code.
Before understanding `if`/`else` statements, we need to review **booleans** - i.e., `TRUE` and `FALSE`

#### Aside: `TRUE` and `FALSE` values

```{r}
TRUE
```

```{r}
20 == 20
```

```{r}
20 > 40
```

A `!` sign can be used as a logical negation

```{r}
!(20 > 40)
```

There are many logical operators to consider.

***

#### Challenge Describe what each fo the following operators are doing.

```{r}
x <- TRUE
y <- FALSE
```

```{r}
x & y
```

```{r}
x | y
```

```{r}
xy <- c(x,y)
```

```{r}
any(xy)
```

```{r}
all(xy)
```

***

Now that we know a bit about booleans, let's get into `if`/`else` statements.

Essentially, an `if` is a conditional that says, "do this thing after the `if` statement, ***if*** the conditional was TRUE."

Here's a simple example (taken from Software Carpentry's lessons)

```{r}
num <- 37
if (num > 100) {
  print("greater than 100")
} else {
  print("not greater than 100")
}
```

#### Challenge 

Re-define `num` so you get the other option.

We don't need an `else` statement for this to work -

```{r}
num <- 37
if (num > 100) {
  print("The number was greater than 100")
}

```

We can also write a "cascade" of `if`/`else` statements

```{r}
if (num > 0) {
  return(1)
} else if (num == 0) {
  return(0)
} else {
  return(-1)
}
```

#### Challenge

Make the above into a function, that takes in any value, and returns whether it is positive, negative, or equal to 0.

***

# Data Visualization and Exploration

The first step to any data analysis is to interrogate your data by calculating some standard statistics and by visualizing it in various ways. 
In future classes we will look more closely at standard statistics.
Today, we'll focus on data visualization.

### Iris dataset

We're going to work with the
[iris dataset](https://stat.ethz.ch/R-manual/R-patched/library/datasets/html/iris.html) again.


```{r}
data( iris )
summary(iris)
```


#### Aside: A review of navigating a `data.frame`

#### Challenge

What commands can I use if I want the first 10 elements of both the sepal width and the species columns?


#### Challenge

What am I going to get if I execute the command below?

```{r, results='hide'}
head(iris[c("Sepal.Width","Petal.Length")])
```

***

## Visualization using `base` R functions

`base` R is how we refer to the things that come pre-loaded in your R environment. Think of them as the **basics**. We'll talk more about add on **packages** in just a bit.

Let's add a habitat type variable to the iris data set. We'll use this later. 
*Caveat - I made up these habitat type preferences.*

```{r}
iris_habitat <- data.frame( Species = c( "setosa", "versicolor", "virginica" ),
                            Habitat = c( "forest", "wetland", "meadow" ) )

iris_full <- merge( x = iris, y = iris_habitat, by = "Species" )

head( iris_full )
tail( iris_full )
```

### Visualizing the measurements of a single variable

Perhaps the most common way to look at data for a single variable is a histogram.
This essentially is a bar plot, where each bar represents the number of times a value falls within a particular *bin*.

Example: let's look at the distribution of `Petal.Length` values

```{r}
hist(iris_full$Petal.Length)
```

***

#### Challenge - use `?hist` to determine how to change the number of bins used

Hint: bins are sometimes called breaks

```{r}
?hist
hist(iris_full$Petal.Length, breaks = 30)
```

***

### Visualizing relationships between two variables

We would like to plot Sepal.Length versus Petal.Length. We'll first do this using
**base R** functions.

```{r plot1}
plot( x = iris_full$Sepal.Length, y = iris_full$Petal.Length )
```

Note, we could save a little typing by using the `with` function. This 
sets up a temporary environment where all of the variables (columns) 
in my data set are defined individually.

```{r plot1-with}
with( data = iris_full, plot( x = Sepal.Length, y = Petal.Length ) )
```


## Visualization using the `ggplot2` package

Now we're going to introduce a data visualization package called **ggplot2**.
This package is great for producing publication quality graphics, but the syntax
used to create a plot is a little more involved.

#### Aside: Installing and loading packages

First, let's install the `ggplot2` package:

```{r installggplot2, eval=FALSE}
# Only need to do this once
install.packages("ggplot2")
```

Then load it:

```{r loadggplot2, message=FALSE}
library(ggplot2)
```


OK, let's fire up ggplot.

```{r plot1-ggplot}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) )
```

Let's break down that call to introduce a few key things about ggplot

* ggplot: the initial canvas we're working on
* geom: geometric objects (i.e. the type of plot - histogram, points, line, etc) 
* aes: aesthetic mapping 

**THAT SEEMS SO COMPLICATED!**

It's true, for simple plots, ggplot can be much more complicated
than simply using the base functions. But the power of ggplot lies
in the ability to lay several geometries (geoms) over each other.
Also, each geometry has a rich set of options. For example,
let's say I want to create the plot we just made, but have 
each species represented by a different color.

```{r plot2}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species ) )
```

Let's add more information - how about habitat type as well.

```{r}
ggplot() + # First make an empty ggplot canvas. Notice the trailing plus sign
  geom_point( data = iris_full, aes( x = Sepal.Length, y = Petal.Length, colour = Species, shape = Habitat), size = 2.5 ) +
  theme_bw()
```


**facets** - a way to separate data into different subplots

Let's say we wanted different plots for each species. We can do that in ggplot using
`facets`.

```{r plot2-facets}
ggplot( data = iris_full, aes( x = Sepal.Length, y = Petal.Length ) ) + 
  geom_point() +
  facet_grid( Species ~ . )
```

**NOTE**: I moved the `data ...` stuff into the initial `ggplot` call here.

*** 

#### Challenge

1. `ggplot2` has many geometries, allowing us to make lot's of different 
types of plots. Let's make two new plots - one **boxplot** of `Petal.Length`, 
with one boxplot for each species. Use `geom_boxplot` for this.

```{r}
ggplot(data = iris_full, aes(x = Species, y = Petal.Length)) +
  geom_boxplot(aes(fill = Species)) +
  theme_bw()  
```

2. Make a histogram of Petal.width. 

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length)) +
  theme_bw()  
```


Use facets to separate the three species.

```{r}
ggplot() +
  geom_histogram(data = iris_full, aes(x = Petal.Length, fill = Species)) +
  facet_grid( Species ~ . ) +
  theme_bw()  
```

3. Second, let's make density plots of Petal.Width. Use `geom_density` and
`colour` to make different colored density lines for iris in each 
habitat type. 
**Note: the area under the curve is equal to 1.**

```{r}
ggplot() +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()
```

4. Use histogram to plot density instead of counts.

```{r}
ggplot() +
  geom_histogram(data = iris_full, 
                 aes(x = Petal.Length, y = ..density.., fill = Species)) +
  facet_grid( Species ~ . ) +
  geom_density(data = iris_full, aes(x = Petal.Length, colour = Species)) +
  theme_bw()  
```


***

### More `ggplot2` resources

* <http://docs.ggplot2.org/current/>: The official **ggplot2** documentation.
* <http://www.amazon.com/o/ASIN/0387981403/ref=nosim/gettgenedone-20>: The **ggplot2** book, by the developer, Hadley Wickham.
* <https://groups.google.com/forum/#!forum/ggplot2>: The **ggplot2** Google Group (mailing list, discussion forum).
* <https://github.com/swcarpentry/bc/tree/master/intermediate/r/data-visualization>: Intermediate Software Carpentry lesson on data visualization with **ggplot2**.
* <http://learnr.wordpress.com/>: A blog with a good number of posts describing how to reproduce various kind of plots using **ggplot2**.
* <http://stackoverflow.com/questions/tagged/ggplot2>: Thousands of questions and answers tagged with "ggplot2" on Stack Overflow, a programming Q&A site.

***
