---
title: "Meeting 3 - In Class Notes"
author: "Matt"
date: "2/12/2020"
output: html_document
---

# Goals

* Continue with intro to R - focusing on programming in R
* Learn about functions
* Learn about for loops and while loops
* Introduce to if and else statements
* Understanding the role of pseudocode

# Functions

## What is a function

R code, grouped together, it does some "thing", and
when we group that code together, we make it generic

## Why write a function

* It makes for cleaner code
* It makes code more understandable
* It's efficient - NO COPY PASTE!

### Example - convert from fahrenheit temperature scale to Kelvin

* substract 32
* multiply by (5/9)
* add 273.15

```{r}
my_temp <- 41
((my_temp - 32) * (5/9)) + 273.15

my_temp2 <- 98.6
((my_temp2 - 32) * (5/9)) + 273.15
```

### Let's write a function

```{r}
fahr_to_kelvin <- function(temp){
  kelvin <- ((temp-32) * (5/9)) + 273.15
  return(kelvin)
}
```

Use our function

```{r}
fahr_to_kelvin(temp = 32)
```

Practice with another function

```{r}
kelvin_to_celsius <- function(temp){
  celsius <- temp - 273.15
  return(celsius)
}
```

Use my function

```{r}
kelvin_to_celsius(273)
```

*** 

#### Challenge

Write a function that converts from Fahrenheit to Celsius

```{r}
farh_to_cels1 <- function(temp){
  celsius <- ((temp-32) * (5/9)) 
  return(celsius)
}

farh_to_cels1(temp = 32)

fahr_to_cels2 <- function(temp){
  as_kelvin <- fahr_to_kelvin(temp = temp)
  as_celsius <- kelvin_to_celsius(temp = as_kelvin)
  return(as_celsius)
}

fahr_to_cels2(32)
```

***

# `for` loops

the answer to the quesiton, how do I do the same thing many times?

```
for(variable in a collection of variables){
  do something to or with that variable
}
```

### Create a temperature data set

```{r}
set.seed(8)
temp_data <- runif(n = 20, min = -5, max = 5) + 45
temp_data
```

Reset our seed to default

```{r}
rm(.Random.seed, envir = globalenv())
```


### For loops


```{r}
for(x in temp_data){
  print(x)
}
```

Another small exampel

```{r}
for(x in 1:10){
  print(x)
}
```

```{r}
for(foo in 10:1){
  print(foo)
}
foo <- as.numeric(5)
```

Let's use a function multiple times

```{r}
for(x in temp_data){
  fahr_to_kelvin(x)
}
```


```{r}
for(x in temp_data){
  print(fahr_to_kelvin(x))
}
```

Use for loop, iteration by element location

```{r}
1:length(temp_data)

for(x in 1:length(temp_data)){
  print(fahr_to_kelvin( temp_data[x] ))
}

```

Loop through by element location, with output as a new vector


```{r}
temp_data_kelvin <- vector()

for(x in 1:length(temp_data)){
  temp_data_kelvin <- c(temp_data_kelvin,
                        fahr_to_kelvin(temp_data[x]))
}

temp_data_kelvin
```

#### Challenge

Use a for loop and write a function to calculate the mean of a vector of numbers.

Mean = (sum a a vector of numbers) / length of that vector

Inside your function, you will use a for loop to calculate the sum of
vector of numbers.

* use a for loop to calculate a sum of numbers
* make a function that includes that for loop so you can do this to
any given set of numbers
* calculate a mean be dividing that sum by the length of those numbers


```{r}
foo <- 1:10
mean(foo)


my_mean <- function(my_vect){
  
  # Calculate the sum of given vector, my_vect
  my_sum <- 0
  for(x in my_vect){
    my_sum <- my_sum + x
    #print(my_sum)
  }
  
  # Divide my sum of the given vector by the length of that vector
  my_calc_mean <- my_sum / length(my_vect)
  
  # Return my_calc_mean
  return(my_calc_mean)
  
}

my_mean(my_vect = foo)
mean(foo)
```

# Conditionals - flow control

`if` `else`

### Booleans

```{r}
TRUE
FALSE
T
F
```

```{r}
20 == 20
```


```
<- does the same thing as =
```

```{r}
bar = 5
bar <- 5
```

```{r}
TRUE == TRUE
TRUE == FALSE
```

```{r}
20 > 40
```

! is logical negation

```{r}
!(20>40)
```

***

#### Challenge

```{r}
x <- TRUE
y <- FALSE
```


```{r}
x & y
```


```{r}
x | y
```

```{r}
xy <- c(x, y)
```

```{r}
any(xy)
```

```{r}
all(xy)
```


```{r}
# Are any of these greater than 5
any(1:10 >= 5)

# Are all of these greater than 5?
all(1:10 >= 5)
```



Finally get to `if`s

```{r}
num <- 37
num > 100

if(num > 100){
  print("greater than 100")
} else {
  print("not greater than 100")
}

```

## Cascade of if/else statements

```{r}
num <- 0

if (num > 0) {
  print("positive")
} else if (num == 0) {
  print("equals 0")
} else {
  print("negative")
}
```

# While loops

```{r}
foo <- 0
while(foo < 10){
  print(foo)
  foo <- foo + 1
}

```

