---
title: "Meeting 5 - In Class Notes"
author: "Matt"
date: "2/26/2020"
output: html_document
---

# Probability through M&Ms

```{r}
mm_counts <- c(57, 55, 56, 55, 55, 43, 54, 56, 50)
```


On average, how many m&ms will I get in a bag

```{r}
sum(mm_counts) / length(mm_counts)
```

Sample mean

$$
\bar{x} = \frac{\sum_{i=1}^n{x_i}}{n}
$$
There is also a population mean, denoted by $\mu$.

```{r}
mean(mm_counts)
```

## Quantify variation in the number of M&Ms we might expect to get in a bag

* Standard deviation - will return to this shortly

* total distance of my observed m&m counts from the mean count

```{r}
mm_counts
dist_from_mean <- mm_counts - mean(mm_counts)
dist_from_mean

```

Get a single number to discuss variability

```{r}
sum(dist_from_mean)
```


* Sum of absolute values

```{r}
sum(abs(dist_from_mean))
```


* sum of squared values

```{r}
sum(dist_from_mean^2)
```

Also called squared deviations from the mean

Advantage of squared vs absolute - squared yields a continuous function, where as absolute has a point of discontinuity.

* variance

sum of squared distances / n-1

$$
var = \frac{\sum_{i=1}^n{(\bar{x} - x_i)^2}}{n-1}
$$

Intuitively think of this as - is the average squared distance of any observation

Take the square root of var, and you get the average absolute distance from the mean of any observation

standard deviation = square root of variance

***

#### Challenge

1. calculate the variance of the number of M&Ms in a bag, using basic functions - `sum` `/` `length`
Check your answer against `var`

2. Do the same for standard deviation.


```{r}
sum( (mean(mm_counts) - mm_counts)^2 ) / (length(mm_counts) - 1)
```

```{r}
var(mm_counts)
```


### do this all for SD

```{r}
sqrt(sum( (mean(mm_counts) - mm_counts)^2 ) / (length(mm_counts) - 1))

```

```{r}
sd(mm_counts)
```

```{r}
library(ggplot2)

ggplot() +
  geom_point(aes(x = mm_counts, y = 0)) +
  geom_vline(xintercept = mean(mm_counts))
```

# Probability

Choose an m&m

You could have chosen a:

{brown, blue, yellow, orange, red, green} a set, our sample space

sample space = is the set that includes all possible outcomes of an event

Event = the selection of one m&m

```{r}
runif(1)
```

Assuming all color m&ms are equally likely,
what's the probability that your event is a green m&m?

1/6


Let's get our data set

```{r}

mm_data <- read.csv(file = "mm_data.csv")
mm_data
```


#### A brief intro into tidyverse

```{r, eval=FALSE}
install.packages("tidyverse")
```

```{r}
library(tidyverse)

mm_data

```


Transform our data into long-format

```{r}
mm_data_long <- tidyr::gather(mm_data, "Person", "MM_Count", 2:10)
mm_data_long

```


**Calc total M&Ms in each bag**

```{r}
mm_data_long %>%
  group_by(Person) %>%
  summarise(Tot_MM = sum(MM_Count))
```

**Add a new column to mm_data_long with the total MM counts**

```{r}
mm_data_long <-
  mm_data_long %>%
  group_by(Person) %>%
  mutate(Tot_MM = sum(MM_Count))

mm_data_long
```

**Calc probability of getting any particular color for each person**

```{r}
mm_data_long$Prob_Color <- mm_data_long$MM_Count / mm_data_long$Tot_MM
mm_data_long
```


***

#### Challenge

What is the man probablity of getting each color across the bags?


```{r}
mm_data_long %>%
  group_by(Color) %>%
  summarise(Prob_Color = mean(Prob_Color))
```

## Probability of events

whats the prob(green OR blue OR red)?

= prob(green) + prob(blue) + prob(red)


what's the prob of not green?

= 1 - p(green) = p(blue) + prob(red) + prob( brown) + p(yellow) + p(orange)

$$
P(\sim Green) = P(Not Green) = P(Green)^c
$$


### Sampling with replacement

I draw one M&M from my bag, put it back, then draw another

* What's the P(green and then blue)?

P(green) * P(blue)

### Samlping without replacement

I draw one M&M, eat it, then draw another

* What's the P(Green and then blue)?

*** 

#### Challenge 

What is the sample space when drawing two M&Ms?

***


```{r}
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")
mm_probs <- c(0.23, 0.14, 0.16, 0.2, 0.13, 0.14)
```

Let's "sample" a new bag

```{r}
new_bag <- sample(x = mm_colors, size = 57, replace = TRUE, prob = mm_probs)
new_bag

table(new_bag)

```

Get new bag color counts

```{r}
new_bag_counts <- c(sum(new_bag == "blue"),
                    sum(new_bag == "brown"),
                    sum(new_bag == "green"),
                    sum(new_bag == "orange"), 
                    sum(new_bag == "red"),
                    sum(new_bag == "yellow")
                    )
new_bag_counts

```


```{r}
mm_matts <- mm_data$Matt
mm_matts
sum(mm_matts)
```

Does my bag match the simulated bag?

```{r}
all(mm_matts == new_bag_counts)

```





Another, side, example of sample

Shuffle numbers / permute numbers

```{r}
sample(1:10, replace = FALSE)
```


Sample w replacement (boot strap)

```{r}
sample(1:10, replace = TRUE)
```




