---
title: "Linear Regression Model Diagnostics"
author: "Matthew Aiello-Lammens"
output: html_document
---

# Flour Beetle Regression

```{r}
flr_beetle <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter8/Data/nelson.csv")

flr_beetle_lm <- lm(data = flr_beetle, WEIGHTLOSS~HUMIDITY)

summary(flr_beetle_lm)
```


Let's use the plot function on the `lm` object.

```{r}
plot(flr_beetle_lm)
```














# Species Abundance vs Area Relationship

From Logan 2010 - Example 8D. Peake and Quinn (1993) investigated the relationship between the abundance of non-mussel invertabrates (`INDIV`) in 25 intertidal mussel clumps, where the clumps covered varying amounts of area (`AREA`), measured in mm^2^. 


```{r}
inverts <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter8/Data/peake.csv")
```

Can we make our data appear normal by transforming them?

Transform = log10

```{r}
library(ggplot2)
ggplot(data = inverts, aes(sample = log10(INDIV))) +
  geom_qq() +
  geom_qq_line()
```


```{r}
ggplot(data = inverts, aes(x = log10(INDIV))) +
  geom_histogram()
```







```{r}
ggplot(data = inverts, aes(sample = log10(AREA))) +
  geom_qq() +
  geom_qq_line()
```

```{r}
inverts_lm_notransform <- lm(data = inverts, INDIV ~ AREA)

summary(inverts_lm_notransform)
```


```{r}
inverts_lm_transform <- lm(data = inverts, log10(INDIV) ~ log10(AREA))

summary(inverts_lm_transform)
```


```{r}
plot(inverts_lm_notransform)
```


```{r}
plot(inverts_lm_transform)
```

