---
title: "Interpreting ANOVA regression coefficients"
author: "Matthew Aiello-Lammens"
output: html_document
---

Let's work with the limpet data set from Quinn (1988), presented in both Quinn and Keough 2009 and Logan 2010. 

```{r}
# Read in the file
limpets <- read.csv(file = "https://mlammens.github.io/ENS-623-Research-Stats/data/Logan_Examples/Chapter12/Data/quinn.csv")
# Make DENSITY a factor
limpets$DENSITY <- as.factor(limpets$DENSITY)
# Show a summary of the data.frame
summary(limpets)
```


Next, let's calculate the group means for the different treatments.

```{r}
library(dplyr)

limpets_grp_means <-
  limpets %>%
  group_by(SEASON, DENSITY) %>%
  summarise(GROUP_Mean = mean(EGGS))
limpets_grp_means

```

Use `spread` function in the `tidyr` package to make this into a nice table

```{r}
library(tidyr)
spread(limpets_grp_means, SEASON, GROUP_Mean)
```

Make a visualization of these data too.

```{r}
library(ggplot2)

ggplot(data = limpets) +
  geom_boxplot(aes(x = DENSITY, y = EGGS, fill = SEASON))
```


Let's now run an ANOVA analysis.

```{r}
limpets_lm <- lm(data = limpets, formula = EGGS ~  SEASON + DENSITY)
anova(limpets_lm)
```

We are confident there is an effect of **both** DENSITY and SEAON on EGGS, based on these results.

Now, we can use the regression coefficients from this model to ***estimate*** the group means.

```{r}
summary(limpets_lm)
```

Let's think of these as regression coefficients (which they are!) and use the following symbols:

|Coef. symbol | Coef. name |
|:------------|:----------:|
|*b~0~*       |(Intercept) |
|*b~1~*       |SEASONsummer|
|*b~2~*       |DENSITY15   |
|*b~3~*       |DENSITY30   |
|*b~4~*       |DENSITY45   |

The table below translates how these coefficients can be used to calcualte *estimates* for each group mean value.

**NOTE** - I'm emphasizing that these are *estimates* of the group means. In this case, our model assumes that there is **no interaction** between the predictor variables DENSITY and SEASON. 

|            |SEASON - Spring| SEASON - Summer |
|:-----------|:-------------:|:---------------:|
|DENSITY - 8 |*b~0~*         |*b~0~* + *b~1~*  |
|DENSITY - 15|*b~0~* + *b~2~*|*b~0~* + *b~1~* + *b~2~*|
|DENSITY - 30|*b~0~* + *b~3~*|*b~0~* + *b~1~* + *b~3~*|
|DENSITY - 45|*b~0~* + *b~4~*|*b~0~* + *b~1~* + *b~4~*|

Let's now use the results of our `lm` call to estimate the group means.
I'm going to do this in the Rmd file by getting the model coefficients, then using the in-line R code execution in the table.


```{r}
limpets_coef <- round(coef(limpets_lm), digits = 3)
```

|            |SEASON - Spring| SEASON - Summer |
|:-----------|:-------------:|:---------------:|
|DENSITY - 8 |`r limpets_coef[1]`|`r limpets_coef[1] + limpets_coef[2]` |
|DENSITY - 15|`r limpets_coef[1] + limpets_coef[3]`|`r limpets_coef[1] + limpets_coef[2] + limpets_coef[3]` |
|DENSITY - 30|`r limpets_coef[1] + limpets_coef[4]`|`r limpets_coef[1] + limpets_coef[2] + limpets_coef[4]` |
|DENSITY - 45|`r limpets_coef[1] + limpets_coef[5]`|`r limpets_coef[1] + limpets_coef[2] + limpets_coef[5]` |

If we compare these estimates to the actual values we see in the table above, we should see that they are pretty close. 

Now, let's see what happens if we include and interaction between DENSITY and SEASON.

```{r}
limpets_lm_int <- lm(data = limpets, EGGS ~ SEASON * DENSITY)
anova(limpets_lm_int)
```

And now let's look at the regression coefficients.

```{r}
summary(limpets_lm_int)
```

As above, let's think of these as regression coefficients (which they are!) and use the following symbols:

|Coef. symbol | Coef. name |
|:------------|:----------:|
|*b~0~*       |(Intercept) |
|*b~1~*       |SEASONsummer|
|*b~2~*       |DENSITY15   |
|*b~3~*       |DENSITY30   |
|*b~4~*       |DENSITY45   |
|*b~5~*       |SEASONsummer:DENSITY15|
|*b~6~*       |SEASONsummer:DENSITY30|
|*b~7~*       |SEASONsummer:DENSITY45|


The table below translates how these coefficients can be used to calcualte *estimates* for each group mean value. 
Pay particular attention to the fact that we now have **three** more coefficients to work with, representing the interaction effects.

|            |SEASON - Spring| SEASON - Summer |
|:-----------|:-------------:|:---------------:|
|DENSITY - 8 |*b~0~*         |*b~0~* + *b~1~*  |
|DENSITY - 15|*b~0~* + *b~2~*|*b~0~* + *b~1~* + *b~2~* + *b~5~*|
|DENSITY - 30|*b~0~* + *b~3~*|*b~0~* + *b~1~* + *b~3~* + *b~6~*|
|DENSITY - 45|*b~0~* + *b~4~*|*b~0~* + *b~1~* + *b~4~* + *b~7~*|

Now using this table, let's use the results of our `lm` call to estimate the group means.

```{r}
limpets_int_coef <- round(coef(limpets_lm_int), digits = 3)
```

|            |SEASON - Spring| SEASON - Summer |
|:-----------|:-------------:|:---------------:|
|DENSITY - 8 |`r limpets_int_coef[1]`|`r limpets_int_coef[1] + limpets_int_coef[2]` |
|DENSITY - 15|`r limpets_int_coef[1] + limpets_int_coef[3]`|`r limpets_int_coef[1] + limpets_int_coef[2] + limpets_int_coef[3] + limpets_int_coef[6]` |
|DENSITY - 30|`r limpets_int_coef[1] + limpets_int_coef[4]`|`r limpets_int_coef[1] + limpets_int_coef[2] + limpets_int_coef[4] + limpets_int_coef[7]` |
|DENSITY - 45|`r limpets_int_coef[1] + limpets_int_coef[5]`|`r limpets_int_coef[1] + limpets_int_coef[2] + limpets_int_coef[5] + limpets_int_coef[8]` |


