---
title: "Assignment 2 - Practicing with `ggplot2`"
output: html_document
---

## Learning Outcomes

* Practice using `ggplot2` to create data visualizations.
* Identify the uses of different types of visualizations.
* Explain the biological or environmental significance of a data visualization.

#### Grading note

* Problem 1 is worth 5 points
* Problem 2 is worth 5 points


## Problem 1 - Plotting relationships between variables

In [section 03](http://mlammens.github.io/ENS-623-Research-Stats/lectures/03-Data-Visualization-and-Exploration.html) we learned to use `ggplot2` to make an x-y scatter plot comparing `Sepal.Length` and `Petal.Length` in the `iris` data set.

Recall that you must first install `ggplot2` if you have not done so already. **ONLY DO THIS IF YOU HAVE NOT ALREADY INSTALLED GGPLOT2**:

```{r, eval=FALSE}
install.packages("ggplot2")
```

Then load it into your environment:

```{r}
library(ggplot2)
```

The code to create one of the plots from section 03 looks like this:

```{r}
data(iris)

ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point()
```

We also learned that you can add the term `geom_smoth(method = "lm")` to add linear regression lines onto your plot, as such:


```{r}
# You will modify this code, as per description below
ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point() +
  geom_smooth(method = "lm")

```

*Use the code above, but replace `Petal.Length` with `Sepal.Width`. Make a new x-y scatter plot with linear regression lines. Describe how having separate lines for each species versus one line for all of the data combined influences your interpretations of these data, and specificially the relationship between `Sepal.Length` and `Sepal.Width`.*

***Hint:** You should make another plot where you **do not** include any `color` arguments to see what happens. That is, a plot where you ignore the species identifications.*

**Put your written answer to the question here.**

## Problem 2 - Histograms

Below is a code chunk to create a histogram of **petal length** values using the `iris` data set.

```{r, eval=FALSE}
ggplot() +
  geom_histogram(data = iris, aes(x = Petal.Length, fill = Species), 
                 position = "dodge") +
  theme_bw() 
```

Modify this code to make a histogram for `Sepal.Length` instead.

```{r}
# Put your code here.
```

*What differences do you notice about the values, as presented in these histograms, for petal length versus sepal length?*


**Put your written answer to the question above, here.**



