---
title: "Possible Problems for Problem Sets"
author: "Matthew Aiello-Lammens"
output: html_document
---

## Blood Oxygen Levels

You are taking measurements of blood oxygen levels in class using a pulse oximeter. Miraculously, you know that the **population mean**, $\mu$, is 97.6 f, and the **population standard deviation**, $\sigma$, is 1.5 f.
Below is a data set of ten measurements taken during class.

```{r}
oxy_data <- data.frame(name = c("Matt", "Yaw", "Madison", "Bryan", "Holly", 
                                "Kevin", "Amanda", "Tyler", "Iris", "Catherine"),
                       blood_ox = c(98.86, 93.16, 98.77, 94.98, 96.03,
                                    98.11, 96.57, 92.88, 94.87, 99.71))
oxy_data
```

### Histogram (10 points)

Make a histogram of these data. 


```{r}
# Put your code here
```

Based on the histogram, what probability distribution do you think best matches your data?

*Put your answer here.*

### Z-scores (10 points)

Assuming that blood oxygen levels are normally distributed, calculate the Z-scores for each of the observations (i.e., each of your classmates).

```{r}
# Put your code to calculate Z-scores here
```

Based on the Z-scores, are there any observations that stand out as being particularly concerning?

*Put your answer here.*


### Identifying Assumptions (5 points)

Identify at least one problem with assuming that these data are normally distributed.

*Put your answer here.*

