---
title: "ENS 623 - Spring 2020 - Problem Set 2"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**
See the introduction to [Problem Set 1](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-1.html) if you need more information about the Rmd format.

#### File name

Your file should use the following naming scheme

**`[last name]_ENS623_SP20_PS2.Rmd`**

For example, 

`Lammens_ENS623_SP20_PS2.Rmd`

#### Grading note

* Problem 1 is worth 15 points
* Problem 2 is worth 10 points


## Problem 1 - Plotting relationships between variables

In class on Wednesday, I demonstrated how to use `ggplot2` to make an x-y scatter plot comparing `Sepal.Length` and `Petal.Length` in the `iris` data set.

Recall that you must first install `ggplot2` if you have not done so already:

```{r, eval=FALSE}
install.packages("ggplot2")
```

Then load it into your environment:

```{r}
library(ggplot2)
```

Then the code to create the actual plot looks like this:

```{r}
data(iris)

ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point()
```

I very briefly showed you that you can add the term `geom_smoth(method = "lm")` to add linear regression lines onto your plot, as such:


```{r}
ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point() +
  geom_smooth(method = "lm")

```

**Use the code above, but replace `Petal.Length` with `Sepal.Width`. Make a new x-y scatter plot with linear regression lines. Describe how fitting a different line to the data from each species might influence how you interpret your data, in comparison to an x-y scatter plot where you do not seperate the species.**

***Hint:* You should make two plots, one where you color the points by species and a second where you do not include any color.**

```{r, eval=FALSE, echo=FALSE}
ggplot(data = iris, aes( x = Sepal.Length, y = Sepal.Width, colour = Species )) + 
  geom_point() +
  geom_smooth(method = "lm")

```


## Problem 2 - Working with syntax in `ggplot2`

Below are two code chunks, both which create histograms and density plots using the `iris` data set.

### Chunk 1

```{r, eval=FALSE}
ggplot() +
  geom_histogram(data = iris, aes(x = Petal.Length, y = ..density.., fill = Species), 
                 position = "dodge") +
  geom_density(data = iris, aes(x = Petal.Length, colour = Species)) +
  theme_bw() 
```

### Chunk 2

```{r, eval=FALSE}
ggplot(data = iris, aes(x = Petal.Length)) +
  geom_histogram( aes(y = ..density.., fill = Species), 
                 position = "dodge") +
  geom_density( aes(colour = Species)) +
  theme_bw() 
```

**Describe the differences between the code chunks.**

**Do Chunks 1 and 2 produce the same plots? (You need to show me that you ran the code to get credit for this.)**

**Based on the answers to the above two questions, what do you think the role of defining arguments in the `ggplot()` function is versus defining arguments in the `geom_` functions?**
