---
title: "ENS 623 - Spring 2021 - Problem Set 1"
output:
  word_document: default
  html_document: default
---

**Please submit this homework on Classes as *both* an R Markdown (Rmd) and DOCX file.**

As I mentioned during our first meeting, rmarkdown is a way to incorporate text, with simple formatting, with R code chunks.
Rather than spell out how to create an Rmd file here, I recommend you consult the following resources:

* [RStudio Rmarkdown Quick Tour](http://rmarkdown.rstudio.com/authoring_quick_tour.html)
* [RStudio Rmarkdown Intro Article](http://rmarkdown.rstudio.com/articles_intro.html)
* [OurCodingClub Intro Example](https://ourcodingclub.github.io/2016/11/24/rmarkdown-1.html)

#### Your file should use the following naming scheme

**`[last name]_ENS623_SP20_PS1.Rmd`**

For example, 

`Lammens_ENS623_SP20_PS1.Rmd`

#### Grading note

* Part 1 is worth 10 points
* Part 2 is worth 15 points

# Part 1 - Introducing the Iris data set

The basic R installation includes a package named `datasets` that contains several different example data sets to use in R. In this problem set, we'll work with the `iris` data set. Using any of the help functions we learned in class, `?` or `help.search`, find the help page for the iris data set. 


```{r}
# Show your code to find information on iris here
```


**Briefly describe what these data are. What is being measured? How many measurements? Etc. Please embed your answer within tese doubl asterisks so your answer is bold-faced.**


# Part 2 - Play with the the `iris` data set

Use the `head`, `tail`, and `summary` functions to investigate the `iris` data set. Use the help functions (or search online) to figure out *how* to use these functions and *what* these functions return. Make at least three observations about the data and/or the data set using these functions.

```{r}
# Put your code here
```


**Put your three observations here.**


