---
title: "ENS 623 - Spring 2021 - Problem Set 6"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**
See the introduction to [Problem Set 1](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-1.html) if you need more information about the Rmd format.

#### File name

Your file should use the following naming scheme

**`[last name]_ENS623_SP21_PS6.Rmd`**

For example, 

`Lammens_ENS623_SP21_PS6.Rmd`


## Working with the YRBSS data set

For this assignment, we are going to use the Youth Risk Behavior Surveillance System (YRBSS) data set.
Let's begin by loading the data into our environment.

```{r}
library(openintro)
data(yrbss)
```

Next, draw a random sample of these data, with sample size equal to 100.

```{r}
yrbss_samp <- yrbss[sample(x = 1:nrow(yrbss), replace = FALSE, size = 100), ]
```


## Problem 1 (5 points)

Calculate the **sample mean** and the **95% Confidence Interval** for **height** based on your sample.


```{r}
# Put your code here
```

## Problem 2 (5 points)

You can calculate the **population mean** for **height** for the `yrbss` data set. 
Does your **95% confidence interval** calculated in Problem 1 *include* the true population mean?


```{r}
# Put code to calculate population mean here
```


**Answer to question here.**


## Problem 3 (5 points)

Using the **population mean** calculated in Problem 2, and the sample mean calculated in Problem 1, calculate the $z$ statistic associated with your sample mean. Is your sample mean significantly different from the population mean? **Explain your response.**


```{r}
# Put code to calculate z statistic and associated probability here

```

**Answer to question here.**

# Problem 4 (10 points)

Using your sample, calculate the **sample means** and **confidence intervals** for **height** for males and females separately. Based on these values, is there a difference between male and female height? **Explain your answer.**


```{r}
# Put your code here
```


**Answer to question here.**


