---
title: "ENS 623 - Spring 2019 - Problem Set 1"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**

As I mentioned during our first meeting, rmarkdown is a way to incorporate text, with simple formatting, with R code chunks.
Rather than spell out how to create an Rmd file here, I recommend you consult the following resources:

* [RStudio Rmarkdown Quick Tour](http://rmarkdown.rstudio.com/authoring_quick_tour.html)
* [RStudio Rmarkdown Intro Article](http://rmarkdown.rstudio.com/articles_intro.html)
* [OurCodingClub Intro Example](https://ourcodingclub.github.io/2016/11/24/rmarkdown-1.html)

#### Your file should use the following naming scheme

**`[last name]_ENS623_SP18_PS1.Rmd`**

For example, 

`Lammens_ENS623_SP18_PS1.Rmd`

#### Grading note

* Part 1 is worth 10 points
* Part 2 is worth 15 points

# Part 1 - More Practice with R

For the first part of this assignment, you will use a learning tool named [swirl](https://swirlstats.com/students.html) to walk through another brief introduction to R.
In order to use `swirl` we will need to download a new "package" to your system.
You can cut and paste the code below to do this.
**IF YOU HAVE ALREADY INSTALLED `swirl` THEN SKIP THE `install.packages()` STEP, BUT DO `library(swirl)`**

```{r, eval=FALSE}
install.packages("swirl")
```

After you've installed `swirl`, execute the following command:

```{r}
library(swirl)
```


Next, we need to download the `swirl` lesson we'll be working with today.
Use the code below.

```{r, eval=FALSE}
swirl::install_course("A_(very)_short_introduction_to_R")
```


### What to do if the swirl repository doesn't seem to be working

Sometimes, the code above doesn't work, so we must first download the file, then open it. 
The file can be found [here](https://github.com/ClaudiaBrauer/A-very-short-introduction-to-R/blob/master/documents/A_(very)_short_introduction_to_R.swc), then once it's downloaded, **run the command below**, and navigating to the *.swc file you just downloaded using the file window that opens up from RStudio.


```{r, eval=FALSE}
swirl::install_course(swc_path = file.choose())
```


## Start your `swirl` lesson.

Start swirl by executing the following command:

```{r, eval=FALSE}
swirl()
```

Navigate to the `A (very) short introduction to R` course and go through Modules 1-3.

## Prove that you did the `swirl` lesson

Recreate **two plots** and the **user defined function** that you created during the `swirl` lesson.
Make sure that you do this in your Rmd file you will submit for this homework.

# Part 2 - Play With a Data Set

In this section of the homework, I want you to acquire a data set, describe it, read it into R, use the `head`, `tail`, and `summary` functions to begin investigating these data, apply **at least** two other functions we used in class, and write a description of what you have found by starting to "play" with your data.

For those of you who already have collected data, this is your chance to play with your data in R. 
For those of you who **do not** have your own data yet, you should acquire a publicly accessible data set.
Based on your presentations last week, I have ideas for each of you. 

One great source of publicly available data is the [DataONE repository](https://www.dataone.org/).
You can use a search phrase (e.g., "deer herbivory") in the **Discover** tool on the main webpage, and go from there. [Data Dryad](https://datadryad.org/) is another good data search engine as well.





