---
title: "ENS 623 - Spring 2022 - Assignment 2"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**
See the introduction to [Problem Set 1](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-1.html) if you need more information about the Rmd format.

#### File name

Your file should use the following naming scheme

**`[last name]_ENS623_SP22_PS2.Rmd`**

For example, 

`Lammens_ENS623_SP22_PS2.Rmd`

#### Grading note

* Problem 1 is worth 5 points
* Problem 2 is worth 5 points


## Problem 1 - Indexing `data.frames`

Using the built-in `iris` data set, use the indexing techniques learned in class to create three new `data.frame`s, each which contains **only** measurements for only one of the species.
That is, one `data.frame` for **setosa**, one for **versicolor**, and one for **virginica**.


```{r}
# Put your code here
```

## Problem 2 - `summary` of your new data frames

Run the `summary` function on each of your new `data.frame`s. 
Compare and contrast the summary stats (e.g. the mean values, or median values, etc.) for each of the data sets.

```{r}
# Put your code here.
```


**Put your compare and contast comments regarding the three data.frames here.**
