---
title: "ENS 623 - Spring 2022 - Assignment 2"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**
See the introduction to [Problem Set 1](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-1.html) if you need more information about the Rmd format.

#### File name

Your file should use the following naming scheme

**`[last name]_ENS623_SP22_PS2.Rmd`**

For example, 

`Lammens_ENS623_SP22_PS2.Rmd`

#### Grading note

* Problem 1 is worth 5 points
* Problem 2 is worth 5 points


## Problem 1 - Plotting relationships between variables

In class we learned to use `ggplot2` to make an x-y scatter plot comparing `Sepal.Length` and `Petal.Length` in the `iris` data set.

Recall that you must first install `ggplot2` if you have not done so already. **ONLY DO THIS IF YOU HAVE NOT ALREADY INSTALLED GGPLOT 2**:

```{r, eval=FALSE}
install.packages("ggplot2")
```

Then load it into your environment:

```{r}
library(ggplot2)
```

Then the code to create the actual plot looks like this:

```{r}
data(iris)

ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point()
```

I very briefly showed you that you can add the term `geom_smoth(method = "lm")` to add linear regression lines onto your plot, as such:


```{r}
# You will modify this code, as per description below
ggplot(data = iris, aes( x = Sepal.Length, y = Petal.Length, colour = Species )) + 
  geom_point() +
  geom_smooth(method = "lm")

```

*Use the code above, but replace `Petal.Length` with `Sepal.Width`. Make a new x-y scatter plot with linear regression lines. Describe how fitting a different line to the data from each species might influence how you interpret your data, in comparison to an x-y scatter plot where you do not separate the species.*

***Hint:** You should make a second plot where you do not include any color arguments to see what happens.*

**Put your written answer to the question above, here.**

## Problem 2 - Histograms

Below is a code chunk to create a histogram of **petal length** values using the `iris` data set.

```{r, eval=FALSE}
ggplot() +
  geom_histogram(data = iris, aes(x = Petal.Length, fill = Species), 
                 position = "dodge") +
  theme_bw() 
```

Modify this code to make a histogram for `Sepal.Length` instead.

```{r}
# Put your code here.
```

*What differences do you notice about the values, as presented in these histograms, for petal length versus sepal length?*


**Put your written answer to the question above, here.**



