---
title: "ENS 623 - Spring 2021 - Problem Set 8"
output: html_document
---


# Overview

In this problem set, you are going to work on comparing groups of data **using ANOVA**, to see if there are significant differences between groups.
We will be using data on Phosphorus levels in the soil at Pace University.
In this problem set, you will be testing for whether there is a difference between **sites**. 

First, let's get the data.


```{r soil_data}
soil_data <- read.csv("https://www.dropbox.com/s/b8vanatf6kejuba/soil_p_data.csv?dl=1")
summary(soil_data)
```

# ANOVA "by hand" (10 points)

**NOTE -** when I say "by hand", what I mean is that you should use R to calculate the group means, grand means, residuals, etc.. Please **do not** do these calculations external from R.

Calculate all of the components to perform an ANOVA and build your ANOVA table.
You might want to consult the [OpenIntro supplementary chatper on ANOVA](https://www.openintro.org/download.php?file=stat_extra_anova_calculations&referrer=/stat/textbook.php) for help or **Table 8.2 in Quinn and Keough 2002** (available in Perusall).

**Put your R code below, and use #s to add comments to keep track of your calculations. YOU MUST INCLUDE COMMENTS TO RECIEVE FULL CREDIT**

```{r}
# Start R code here


```



Determine if we reject or accept our null hypothesis, where $H_0$ = there is no difference in phosphorus levels at the different sites. Explain your answer.

**Write your answer here.**


# ANOVA with R (5 points)

Use either `summary(aov())` or `anova(lm())` to perform an ANOVA identical to what you did "by hand" above.

**Put your R code below.**

```{r}



```

Do the results of your "by hand" ANOVA match with the results of using `anova` or `aov`? Explain how you determined that they do or don't match.

**Write your answer here.**


# Tukey's Honest Significant Differences test (5 points)

Use `TukeyHSD()` to perform all multiple comparisons, and to determine with sites are different from each other.

**Put your R code below.**

```{r}



```

Based on the results of the Tukey HSD test, which sites do you think are different from one another? Explain your answer.

**Write your answer here.**

# Results and Discussion (5 points)

Based on all of the analyses you have done here, write a brief results and discussion section. This section should include *at least* three specific observations of your results and at least two potential interpretations of these results that are supported by your analyses.


**Write your answer here.**


