---
title: "ENS 623 - Spring 2019 - Problem Set 3"
output: html_document
---

**Please submit this homework as an R Markdown (Rmd) file.**
See the introduction to [Problem Set 1](http://mlammens.github.io/ENS-623-Research-Stats/problem_sets/Problem-set-1.html) if you need more information about the Rmd format.

#### File name

Your file should use the following naming scheme

**`[last name]_ENS623_SP19_PS3.Rmd`**

For example, 

`Lammens_ENS623_SP19_PS3.Rmd`

#### Grading note

* Problem 1 is worth 10 points
* Problem 2 is worth 15 points

# Problem 1 - How many coins must I flip?

### `while` loops

`while` loops are not necessarily as common as `for` loops introduced in class Wednesday, but they are quite important. 
One key difference between `for` loops and `while` loops is that the former is usually set to repeat a given number of times, while the later repeats until some condition is met.
Incidentally, this also means it's easy to get into an [infinite loop](https://en.wikipedia.org/wiki/Infinite_loop) when writing `while` loops. Be careful!

Here is a basic `while` loop.

```{r}
foo <- 0
while(foo < 10){
  print(foo)
  foo <- foo + 1
}
```

Imagine you are flipping a fair coin (i.e., there is a 0.5 chance of getting a heads and a 0.5 chance of getting a tails).
You want to continue flipping the coin *until* you get 100 heads.
**Write a script that simulates coin flips, and counts the number of times you need to flip the coin until you get 100 heads.**
Note that since the coin flip is a random process, it will not necessarily take you the same number of flips if you re-run the script.

# Problem 2 - How likely was my bag of M&Ms?

Write a short R script to simulate the combinations of colors that would have been possible from your bag of M&Ms assuming the company-stated color distribution (23% blue, 14% brown, 16% green, 20% orange, 13% red, 14% yellow) and use this script to calculate the probability of obtaining the combination in your bag. 

Needed:
* `for` loop
* `if` statement
* `sample` function

Recall that you can sample a bag of M&Ms using the following:

```{r}
## Colors as a vector
mm_colors <- c("blue", "brown", "green", "orange", "red", "yellow")

## Proportion/probability of each color
mm_probs <- c(.23, .14, .16, .20, .13, .14)

## I want to "sample" a bag of MMs
new_bag <- sample(x = mm_colors, size = 15, replace = TRUE, prob = mm_probs)
```

## Hints and tips

* You can read in your data from Wednesday's class in order to use **one** of your bags from class.

* Your bag is pretty unlikely, so it may take **many** iterations in your `for` loop.

* You will need to have a **counter** of some sort, to count the number of times your original bag matches the random bag.

* Make sure you have the M&M counts in the order for the original bag and the new bag.

